---
title: "STT 301 Learning R Markdown - Part 1"
author: "Shawn Santo"
date: "September 10, 2018"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE, comment = NA)
```


## Introduction

---

**Learning objectives:**

  + R markdown syntax
  + Data basics
  + Object assignments
  + Vectors

---

An R markdown (.Rmd) file will allow you to generate reproducible documents
that include your R code, comments, and explanations. All of your homework
assignments and majority of the in-class assignments will be in the form of an
R markdown document.

This .Rmd file demonstrates some of the syntax and style options. 

## Basics

- Everything we'll do comes down to applying **functions** to **data**.

- **Data**:  things like 7, "seven", $7.000$, the matrix 
  $\left[ \begin{array}{ccc} 7 & 7 & 7 \\ 7 & 7 & 7\end{array}\right]$, etc.

- **Functions**: things like $\log{}$, $+$, $<$, $\mod{}$, etc.

A function is a machine which turns input objects (**arguments**) into an
output object (**return value**), possibly with **side effects**, according 
to a definite rule.

## Data building blocks

You'll encounter different kinds of data types:

- **Booleans**: Direct binary values: `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero)
- **Characters**: fixed-length blocks of bits, with special coding;
  **strings** = sequences of characters
- **Floating point numbers**: a fraction (with a finite number of bits) 
  times an exponent, like $1.87 \times {10}^{6}$
- **Missing or ill-defined values**: `NA`, `NaN`, etc.


## Arithmetic operators and some built-in functions

You can use R as a fancy calculator.

Command | Description
--------|-------------
`+,-,*,\` | add, subtract, multiply, divide
`^` | raise to the power of
`%%` | remainder after division (ex: `8 %% 3 = 2`)
`( )` | change the order of operations
`log(), exp()` | logarithms and exponential (ex: `log(10) = 2.302`)
`sqrt()` | square root
`round()` | round to the nearest whole number (ex: `round(2.3) = 2`)
`floor(), ceiling()` | round down, round up 
`abs()` | absolute value

Give an example using each of the commands above. Place these examples inside
the below `R` chunk. Provide a comment as to which command is being used 
(to comment in the script, console, or inside a chunk use #). Below is an 
example using addition.

Multiple lines of code can be placed in a single `R` chunk.


```{r}
10 + 7 # addition

8 - 6 # subtraction

4 * 3 # multiplication

3 / 1 # division

4 ^ 13 # power

10 %% 3 # mod

(4 + 5) / 9

log(exp(5)) # log/exp

sqrt(100) # square root

round(3.14321) # round

floor(43.12) # floor

ceiling(10.1) # ceiling

abs(-4) # absolute value
```


## Operators continued

**Comparisons** are binary operators; they take two objects, like numbers, and 
give a boolean (TRUE or FALSE) response.
  
  Command | Description
--------|-------------
`>`  | greater than
`<`  | less than
`>=` | greater than or equal to
`<=` | less than or equal to
`==` | equal to (recall that `=` is for assignment and not checking equality)
`!=` | not equal to
`&`  | and (ex: `(5 > 7) & (6*7 == 42)` will return the value FALSE)
`|`  | or  (ex: `(5 > 7) | (6*7 == 42)` will return the value TRUE)


Give an example using each of the commands above. These should be inside the
below `R` chunk. Provide a comment as to which command is being used. 
You do not need to make all the comparisons be TRUE.

```{r}
100 > 22 # greater than

-1 < -10 # less than

4 >= 10 # greater than or equal to

pi <= 3 # less than or equal to

1 == "1" # check equality

1 != 1 # not equal

(4 > 1) & (5 > sqrt(10)) # and

(exp(pi) < 100) | (0 > 1) # or

```


## Variables and objects
  
We can assign names to objects. `R` has some built in object names such as `pi`.
```{r}
pi
```

Variables can be arguments to functions or operators, just like constants:
```{r}
pi*10
cos(pi)
```

## Assignment operator

Objects and variables are created with the **assignment operator**, 
`<-` or `=`. Remeber that in RStudio you can use the shortcut `Alt + -` to 
create the `<-` assignment operator. The shortcut includes convenient spacing
before and after the assignment operator. 

```{r}
time_factor <- 12
time_factor
time_in_years = 2.5
time_in_years * time_factor
```

The assignment operator also changes values:
```{r}
time_in_months <- time_in_years * time_factor
time_in_months
time_in_months <- 45
time_in_months
```

Using names and variables makes code easier to design, easier to debug, 
less prone to bugs, easier to improve,  and easier for others to read.

Use descriptive variable names

- Good: `num_students <- 35`
- Bad: `ns <- 35 `

##  The workspace

What names have you defined values for? These will also appear in the
Environment tab within RStudio.

```{r}
ls()
```

If you want to remove a variable use `rm()`.

```{r}
rm("time_in_months")
ls()
```

To remove all objects from your workspace,
```{r}
rm(list=ls()) # remove all variables
```
This is sometimes helpful at the start of a new `R` session.

## Some built-in functions on vectors

Command | Description
--------|------------
`sum(vec)` | sums up all the elements of `vec`
`mean(vec)` | mean of `vec`
`median(vec)` | median of `vec`
`min(vec), max(vec)` | the smallest or largest element of `vec`
`sd(vec), var(vec)` | the standard deviation and variance of `vec`
`length(vec)` | the number of elements in `vec`
`sort(vec)` | returns the `vec` in sorted order
`order(vec)` | returns the index that sorts the vector `vec`
`unique(vec)` | lists the unique elements of `vec`
`summary(vec)` | gives a five-number summary  


Create a vector, and apply three of the functions above. Your vector should have
at least 6 numbers. To create a vector you can use the concatenate function, 
`c()`, with a single comma between each value. To practice good code syntax,
after every comma there should be a space. Be sure to provide a comment as to 
which function you are using.

```{r}
x <- c(1:10)

sum(x) # sum

sd(x) # standard deviation

summary(x) # 5 number summary
```

Below is an example using some of the above functions on a vector named 
`course_grades`.

```{r}
course_grades <- c(80, 90, 93, 82, 95)

mean(course_grades) # compute mean grade
median(course_grades) # compute median grade
sd(course_grades) # grade standard deviation
```


```{r}
sort(course_grades) # type ?sort in your console to see the default sorting technique
max(course_grades) # highest course grade
min(course_grades) # lowest course grade
```

## Referencing elements of vectors
```{r}  
students <- c("Matt", "Louisa", "Frank", "Farhad", "Li")
```
  
Vector of indices
```{r}
students[c(2,4)]
```

Vector of negative indices
```{r}
students[c(-1,-3)]
```

## More referencing

`which()` returns the `TRUE` indices of a boolean vector.
```{r}
course_grades
a_threshold <- 90 # A grade = 90% or higher
course_grades >= a_threshold # vector of booleans
a_students <- which(course_grades >= a_threshold) # Applying which() 
a_students
students[a_students] # Names of A students
```

## Named components
  
You can give names to elements or components of vectors.
```{r}
students
names(course_grades) <- students # Assign names to the grades
names(course_grades)
course_grades[c("Matt","Frank","Li")] # Get final grades for 3 students
```
The labels in what `R` prints are not part of the value.
