---
title: 'STT 301'
author: "Shawn Santo"
date: "September 5, 2018"
output: html_document
---

```{r include=FALSE}
library(knitr)
load("baseball.Rdata")
```


## Introduction

---

**Learning objectives:**

  + Functions for descriptive statistics
  + Variable assignment
  + Data subsetting

---

The `Lahman` package in `R` contains a plethora of baseball data. The package 
is used extensively in the book, "Analyzing Baseball Data with R", by Max Marchi
and Jim Albert (2013). This tutorial will use a subset of data from the 
`Lahman` package to expose you to some basic descriptive statistical functions
and data subsetting.

The subset of data from the `Lahman` package is automatically loaded at the
start of this tutorial. You will learn how to read in different data files later.

## Data analysis basics - part 1

### Accessing data

To see all the objects loaded from the baseball data set in this tutorial 
use the function `ls()`. Function `ls()` lists all the objects in the current
environment. However, since only a subset of the `Lahman` package is loaded,
`ls()` will only show those objects corresponding to the baseball data set.

```{r}
ls()
```

To access the content of an object in `R` use the object's name.
Keep in mind that `R` is case sensitive. Thus, we need to type an object's name
exactly as it appears.

Above we see the object `JimRiceBA`. Run the code below to see the contents of
`JimRiceBA`.

```{r}
JimRiceBA
```

### Descriptive statistics

The names of many functions in `R` are self-explanatory. To compute the minimum,
maximum, and mean for Jim Rice's career batting average we can use the
corresponding functions given below.


```{r}
min(JimRiceBA)
max(JimRiceBA)
mean(JimRiceBA)
```

To find the year in which Jim Rice had his lowest batting average and
the year in which he had his highest batting average,
we can make use of the functions `which.min()` and `which.max()`, respectively.

```{r}
which.min(JimRiceBA)
which.max(JimRiceBA)
```

Let's examine how Rice's batting average changed throughout his career. First, we
compute year-over-year differences, then view the results. Second, we will look at
which year he had the largest increase and which year he had the largest decrease.

```{r}
# compute differences
JimRiceBA_diffs <- diff(JimRiceBA, lag = 1)
JimRiceBA_diffs

# find years
which.max(JimRiceBA_diffs)
which.min(JimRiceBA_diffs)
```

The `#` symbol was used to add comments. `R` does not execute anything 
following `#`. Use `#` for code documentation to explain to
others why you are doing what you are doing. Good code documentation is also 
beneficial for your future self.

## Data analysis basics - part 2

### Summary statistics with two variables

Recall that the correlation measures the linear strength between two
quantitative variables. Let's look at the correlation between each pair of
available variables for Jim Rice: batting average, home runs, and RBIs.

```{r}
cor(JimRiceBA, JimRiceHR)
cor(JimRiceBA, JimRiceRBI)
cor(JimRiceHR, JimRiceRBI)
```

To view a simple plot of Rice's home runs versus his RBIs we can use the
`plot()` function.

```{r}
plot(JimRiceHR, JimRiceRBI)
```

To learn more about Jim Rice click [here](https://en.wikipedia.org/wiki/Jim_Rice).

## Exercise 1

In the following exercises you use the data on [Ted Williams](https://en.wikipedia.org/wiki/Ted_Williams). 

To remind yourself of the variable names use the function `ls()`.

### Part a

Use the `length()` function to determine how many seasons Ted Williams played.

```{r}
length(TedWilliamsBA)
```

### Part b

What season did Ted Williams have his highest batting average?

```{r}
which.max(TedWilliamsBA)
```


### Part c

What was Williams' highest batting average?

```{r}
max(TedWilliamsBA)
```


### Part d

What was Williams' career mean batting average?

```{r}
mean(TedWilliamsBA)
```


### Part e

What was the correlation between Williams' home runs and RBIs? Was it higher than
Jim Rice's correlation?

```{r}
cor(TedWilliamsHR, TedWilliamsRBI)
cor(JimRiceHR, JimRiceRBI)
```
Jim Rice had a high correlation between home runs and RBIs.

### Part f

What was the largest absolute change in Williams' RBIs year-over-year?

```{r}
tw_rbi_diff <- diff(TedWilliamsRBI, lag = 1)
max(abs(tw_rbi_diff))
```


## Data frames

### Subsetting

The `R` objects such as `JimRiceBA` and `TedWilliamsHR` are known as vectors.
For now, think of a vector as data of the same form corresponding to one variable.
In the following section and exercises you will work with data frames. Think of
a data frame as table or a data set you would view in an excel spreadsheet. 
We will formally learn about vectors, data frames, and other data types of `R` 
later.

Among the objects loaded at the start of this tutorial was the data frame
`batting_stats`. Each column of the data frame corresponds to a variable. Let's
see how many variables are in `batting_stats` as well as how many rows the data
frame contains.

```{r}
names(batting_stats)
dim(batting_stats)
```

We see that `batting_stats` has 97,889 rows and 32 variables, where the variable
names are provided by the `names()` function.

One way to subset a data frame is to use `[ , ]` and specify row and column
indices.

```{r}
batting_stats[90000, c(32, 25, 13, 14)]
batting_stats[c(90000, 91000, 92000), c(32, 25, 13, 14)]
```

A more informative and efficient way to subset a data frame is to use the 
`subset()` function. Let's look at Jimmy Rollin's batting average, home runs,
and RBIs.

```{r}
jroll <- subset(batting_stats, playerName == "Jimmy Rollins")
jroll[, c(32, 25, 13, 14)]
```

Suppose we are interested in subsetting `batting_stats` based on the players 
with a batting average greater than 0.400 and who played in a minimum of
100 games.

```{r}
over400 <- subset(batting_stats, BA > 0.400 & G >= 100)
over400[, c("playerName", "BA", "G", "yearID")]
```

Based on the past three examples carefully note:

  1. columns can be subset by the column number or name,
  2. not specifying an index before the comma in `[ , ]` returns all rows,
  3. not specifying an index after the comma in `[ , ]` returns all columns,
  4. to specify a variable is equal to a given value use `==` and not `=`.
  
## Exercise 2

You will use the `batting_stats` data frame to answer the questions that follow.

### Part a

  i. Extract all the data corresponding to Lou Gehrig, and assign the subsetted 
     data set to an object named `lg`.
  ii. Based on the `lg` data frame, display the player's name, home runs, 
      batting average, RBIs, and year.
  iii. Display the same data as in (ii), but only for the years with at least 30
       home runs.

```{r}
lg <- subset(batting_stats, playerName == "Lou Gehrig")
lg_lite <- lg[, c("playerName", "HR", "BA", "RBI", "yearID")]
lg_lite

subset(lg_lite, HR >= 30)
```


### Part b

  i. Extract all the rows from `batting_stats` where the number of home runs is 
     at least 50 and the batting average is above 0.300. Assign this subsetted
     data to an object named `power_plus_average`.
  ii. Which players achieved this feat?
  iii. In what years did these achievements occur?
  
```{r}
power_plus_average <- subset(batting_stats, HR >= 50 & BA > 0.300)
power_plus_average[, "playerName"]
power_plus_average[, "yearID"]
```

  
### Part c

How many players hit more than 80 home runs in a single season?

```{r}
subset(batting_stats, HR > 80)
```


### Part d

What year did Barry Bonds hit a record-breaking 73 home runs for the Giants?

```{r}
bonds <- subset(batting_stats, playerName == "Barry Bonds" & HR == 73)
bonds[, "yearID"]
```


