---
title: "STT 301: Loops, if-else, and functions"
author: "Shawn Santo"
date: "September 26, 2018"
output: html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = NA, tidy = F)
```

## Introduction

---

**Learning objectives:**

  + For, while, repeat loops
  + Writing functions

---

The Fibonacci sequence is the sequence $\{F_n\colon n \geq 1\}$ of numbers
defined by $F_1 = 0, F_2 = 1$ and $F_n = F_{n-2} + F_{n-1}$ for $n\geq 3$. 
For example, the first seven Fibonacci numbers are 
$F_1 = 0, F_2 = 1$, $F_3 = 0+1 = 1$, $F_4 = 1 + 1 = 2$, 
$F_5 = 1 + 2 = 3$, $F_6 = 2 + 3 = 5$, and $F_7 = 3 + 5 = 8$. 

## For, while, repeat loops

### Part a

Write a function, `Fib1`, which uses a `for` loop. `Fib1` should take one 
argument, `n`, and return the first `n` Fibonacci numbers. 
For example, `Fib1(7)` should return the values `0, 1, 1, 2, 3, 5, 8`. 
Test your function by checking the result of `Fib1(7)` and `Fib1(10)`.

```{r}
Fib1 <- function(n) {
  # check for n to be greater than or equal to 1
  stopifnot(n >= 1)
  
  fib_vec <- c(0, 1)
  
  if (n <= 2) { # n = 1 or n = 2
    fib_vec <- fib_vec[1:n]
  }else{
    for (i in 3:n) { # n >= 3
      fib_vec[i] <- fib_vec[i - 2] + fib_vec[i - 1]
    }
  }
  
  return(fib_vec)
}

Fib1(7)
Fib1(10)
```

Did you make sure your function would return the correct value(s) if 
called with `Fib1(1)` or `Fib1(2)`? 

### Part b

Verify below that `Fib1(1)` and `Fib1(2)` return the correct values. 
Modify your function if the values are not correct.

```{r}
Fib1(1)
Fib1(2)
```

### Part c

Write a function called `Fib2` that does the exact same task as `Fib1`, 
but uses a `while` loop as opposed to a `for` loop. 
Test your function by checking the result of `Fib1(1)`, `Fib1(2)`, 
`Fib1(7)`, and `Fib1(10)`.

```{r}
Fib2 <- function(n) {
  # check for n to be greater than or equal to 1
  stopifnot(n >= 1)
  
  fib_vec <- c(0, 1)
  
  if (n <= 2) { # n = 1 or n = 2
    fib_vec <- fib_vec[1:n]
  }else{
    # n >= 3
    i <- 3
    while (i <= n) {
      fib_vec[i] <- fib_vec[i - 2] + fib_vec[i - 1]
      i <- i + 1
    }
  }
  
  return(fib_vec)
}

Fib2(1)
Fib2(2)
Fib2(7)
Fib2(10)
```

### Part d

Write a function called `Fib3` that does the exact same task as `Fib1`, 
but uses a `repeat` loop as opposed to a `for` loop. 
Test your function by checking the result of `Fib1(1)`, `Fib1(2)`, 
`Fib1(5)`, and `Fib1(10)`.

```{r}
Fib3 <- function(n) {
  # check for n to be greater than or equal to 1
  stopifnot(n >= 1)
  
  fib_vec <- c(0, 1)
  
  if (n <= 2) { # n = 1 or n = 2
    fib_vec <- fib_vec[1:n]
  }else{
    # n >= 3
    i <- 3
    repeat {
      fib_vec[i] <- fib_vec[i - 2] + fib_vec[i - 1]
      i <- i + 1
      if (i > n) {break}
    }
  }
  
  return(fib_vec)
}

Fib3(1)
Fib3(2)
Fib3(7)
Fib3(10)
```

## More loops

Write a function called `find_cos_integer` which takes a value 
$x$ between $-1$ and $1$ (inclusive) as an input, and it returns the smallest
nonnegative integer $k$ satisfying the inequality $|\cos(k) - x| < 0.001$. 
Here are a few examples:

`> find_cos_integer(1)`    
`[1] 0`  

`> find_cos_integer(0.1)`  
`[1] 7202`

`> find_cos_integer(0.3)`  
`[1] 49`

Try to include code which stops and returns a message if the user specifies 
an argument $x$ which is either less than $-1$ or greater than $1$. 
Failure to do this may result in an infinite loop.

```{r}
find_cos_integer <- function(x){
  
  k <- 0
  repeat {
    if (abs(cos(k) - x) < 0.001) {break}
    k <- k + 1
  }
  
  return(k)
}

find_cos_integer(0.1)
find_cos_integer(-0.3)
find_cos_integer(0.5)
find_cos_integer(0)
find_cos_integer(1)
```