---
title: "STT 301: Apply family of functions"
author: "Shawn Santo"
date: "October 15, 2018"
output: html_document
---


```{r global_options, echo = FALSE}
knitr::opts_chunk$set(fig.path='Figs/', warning=FALSE, 
                      message=FALSE, dev = "png", comment = NA)
```

## Introduction

---

**Learning objectives:**

- reading data into R
    - `read.csv`
    
  - `apply` function
    - matrices
    - data frames
    - arrays
    
  - apply family

---

```{r echo = TRUE}
library(ggplot2)
```

## The apply function

#### Part a

Download the csv file `nevada_casino_sqft.csv` from D2L. The data set contains Nevada casino square footage by activity. Before you read the csv file into R, open 
it and look at how `NA` values are represented as well as if variable headers 
are present. Then, use `read.csv` to read the file in to R 
(make use of the `na.strings` argument). Save the data frame as 
`casino.na`. Examine the structure of the data frame `casino.na`. 

```{r echo = T}
casino.na <- read.csv(file = "nevada_casino_sqft.csv", 
                      header = TRUE, na.strings = " ", stringsAsFactors = F)
str(casino.na)
```

#### Part b

Use the `apply` function to 

  1. check how many `NA` values are in each row;
  2. find the mean square footage for "PITGAMES", "SLOTS", "KENO", "BINGO", 
     "SPORTS", "POKER", and "TOTAL";
  3. find the median square footage for "PITGAMES", "SLOTS", "KENO", "BINGO", 
     "SPORTS", "POKER", and "TOTAL";
  4. compute the five number summary (use the function `summary`) 
     for "PITGAMES", "SLOTS", "KENO", "BINGO", "SPORTS", "POKER", and "TOTAL";

```{r echo = T}
apply(is.na(casino.na), 1, sum)

apply(casino.na[,4:10], 2, mean, na.rm = TRUE)

apply(casino.na[,4:10], 2, median, na.rm = TRUE)

apply(casino.na[,4:10], 2, summary)

```

#### Part c

Download the same data set (this does not have `NA` values) from "http://www.stat.ufl.edu/~winner/data/nevada_casino_sqft.csv". 
Use this URL directly in the `read.csv` function. Set `stringsAsFactors = FALSE`.
Save the data frame as `casino`. Examine the structure of `casino`.

```{r echo = FALSE}
casino <- read.csv(file = "http://www.stat.ufl.edu/~winner/data/nevada_casino_sqft.csv", header = TRUE, stringsAsFactors = FALSE)
str(casino)
```

#### Part d

Use the `with` function to

  1. create a table for the variable `AREA`;
  2. compute the mean `SPORTS` square footage for when the `SPORTS` square 
     footage is greater than 0;
  3. compute the median `POKER` square footage for when both the `POKER`
     and `SPORTS` square footage are greater than 0.

```{r echo = T}

with(casino, table(AREA))

with(casino, mean(SPORTS[SPORTS > 0]))

with(casino, median(POKER[POKER > 0 & SPORTS > 0]))

```

#### Part e

Write a function that computes something simple. 
Use `apply` to apply your written function to 
columns 4 through 10 of the `casino` data frame.

```{r}
top <- function(x){
  sort(x, decreasing = T)[1:3]
}

apply(casino[, 4:10], 2, top)
```


*************************

## Central Limit Theorem and the apply function

Here you will write a function and then use the `apply` function so you can 
demonstrate the Central Limit Theorem through a simulation.  Assume our 
population distribution is the Poisson distribution. 
The goal is for your function to create similar plots as you see below. 
Thus, the sampling distribution of the sample mean should become more
mound-shaped and symmetric as the sample size increases.

Some helpful hints for writing your function:

  - have your function take two arguments: sample size -`n` and 
    Poisson parameter - `lambda`
  - use `rpois` to generate random variables from the Poisson distribution
  - use 10000 replications for the user specified sample size 
    (thus the function will need to generate n * 10000 random variables)
  - store the generated Poisson random variables in an n x 10000 matrix 
    (use the R function `matrix` and don't forget to specify the matrix dimensions)
  - use `apply` to compute the mean of each column of the matrix
  - plot the distribution of the sample means using the `geom_bar` 
    geom function from `ggplot2`
  - use `set.seed(1360)` inside your function, before you generate the random
    variables, to have your plots match exactly to the below plots


```{r echo = T}

pois.clt <- function(n, lambda){
  set.seed(1360)
  reps <- 10000
  x.mat <- matrix(rpois(n*reps, lambda), n, reps)

  x.means <- apply(x.mat, 2, mean)
  
  x.means.df <- as.data.frame(x.means)
  
  ggplot(data = x.means.df, aes(x = x.means)) + 
    geom_bar(fill = "purple", color = "black") + 
    labs(x = expression(bar(x)), 
         title = expression(paste("Sampling Distribution of ", bar(X),sep="")), 
         subtitle = paste("n = ", n, sep =""))
}
```


```{r echo = T}
pois.clt(n = 1, lambda = 1)

pois.clt(n = 2, lambda = 1)

pois.clt(n = 5, lambda = 1)

pois.clt(n = 10, lambda = 1)

pois.clt(n = 30, lambda = 1)

pois.clt(n = 50, lambda = 1)

```

*************************

## Apply with arrays

Understand each of the below examples using `apply` on a 
three dimensional array. Run the code line-by-line to see what happens.

```{r results="hide"}
# create a 2 x 2 x 3 array that contains the numbers 1 - 12

my.array <- array(data = c(1:12), dim = c(2,2,3))

# view the array

my.array

# apply sum over 1 dimension

apply(my.array, 1, sum)
apply(my.array, 2, sum)
apply(my.array, 3, sum)

# apply sum over multiple dimensions

apply(my.array, c(1,2), sum)
apply(my.array, c(1,3), sum)
apply(my.array, c(2,3), sum)
apply(my.array, c(3,1), sum)
apply(my.array, c(3,2), sum)

```

## A summary of the `a,l,s,t apply` functions

Command | Description
--------|--------------------------------------------
`apply(X, MARGIN, FUN, ...)` | Obtain a vector/array/list by applying `FUN` along the specified `MARGIN` of an array or matrix `X`
`lapply(X, FUN, ...)` | Obtain a list by applying `FUN` to the elements of a list `X`
`sapply(X, FUN, ...)` | Simplified version of `lapply`.  Returns a vector/array instead of list.
`tapply(X, INDEX, FUN, ...)` | Obtain a table by applying `FUN` to each combination of the factors given in `INDEX`  

- these functions are good alternatives to loops

- they are typically more efficient than loops (often run considerably faster on large data sets)

- take practice to get used to, but make analysis easier to debug and less prone to error when used effectively

- you can always type `example(function)` to get code examples (E.g., `example(apply)`)



