---
title: "STT 301: Text Data"
author: "Shawn Santo"
date: "November 19, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.path='Figs/', warning=FALSE, message=FALSE, dev = "png", comment = NA)
```

### Introduction

Learning Objectives
  - string/character manipulation functions
  - regular expressions
  
*************************

### Number of characters

To determine the number of characters in a string use `nchar` and not the `length` function. How many characters are in the following?

  1. "coffee"
  2. "supercalifragilisticexpialidocious"
  3. "what is the character count here?"
  4. "whatisthecharactercounthere?"

```{r}
nchar("coffee")
nchar("supercalifragilisticexpialidocious")
nchar("what is the character count here?")
nchar("whatisthecharactercounthere?")
```

`nchar` can also find the number of characters in each element of a vector. Use `nchar` on a character vector of your choice.

```{r}
colors <- c("red", "blue", "orange", "purple")
nchar(colors)
```


### Getting a substring

```{r}
phrase <- "extract a substring from this phrase"

substr(x = phrase, start = 1, stop = 7)

substr(phrase, 1, 8)

substr(phrase, 5, 20)

```

`substr` also works on a vector. 

```{r}
presidents <- c("Clinton", "Bush", "Regan", "Carter")

# extract the first two letters from each word

substr(presidents, 1, 2)

# extract the last two letters from each word

substr(presidents, nchar(presidents) - 1, nchar(presidents))

# we can grab the first, first 2, first 3, etc

substr(x = presidents, start = 1, stop = 1:4)
```

`substr` also allows us to replace characters.

```{r}
substr(phrase, 1, 2) <- "at"
phrase

# try and replace attract with "grab"

substr(phrase, 1, 7) <- "grab"
phrase

```

### Splitting strings

`strsplit` allows you to separate a string. Do you remember the structure that is returned? For now, use the argument `fixed = TRUE` and do not use regular expressions.

```{r}
big.cats <- "lion, tiger, jaguar, cougar, leopard, snow leopard, cheetah"

# split big.cats at each comma 

strsplit(x = big.cats, split = ",", fixed = TRUE)

# split big.cats at each comma, but make sure the result is a vector

unlist(strsplit(x = big.cats, split = ",", fixed = TRUE))

# split big.cats at each space

unlist(strsplit(big.cats, split = " ", fixed = TRUE))

small.cats <- "serval, caracal"

# split the vector c(big.cats, small.cats) at each comma

strsplit(c(big.cats, small.cats), split = ",", fixed = TRUE)

# we can also split by each character

strsplit(x = small.cats, split = "")

# split the string small.cats at each "a"

strsplit(small.cats, split = "a", fixed = TRUE)

```

### Regular expression (Regex)

The true power of `strsplit` or `grep` is realized when we use a regular expression or a regex.

- A **regular expression** or **regex** is a specially structured string that allow us to match certain patterns occurring in text.
- Note: regexes follow a well-defined set of rules, independent of the R language.
- Any string defines a valid regex. To get us started, we'll consider **literals**, which are just strings that we want to match, literally.
    - "fly" matches "superfly", "why walk when you can fly"  
    - "fly" does not match "time flies like an arrow", "fruit flies like bananas"
- OR of two regexes is a regex.
    - "fly|flies" tries to match "fly" or "flies" 
- Concatenation of regexes is a regex.
    - "(time|fruit) (fly|flies)" tries to match "time" or "fruit", then a space, then "fly" or "flies"
- Parentheses define groups

To match a regular expression or regex we can use the `grep` function. Recall, that `grepl` returns a logical value instead of a location.

```{r}
str.vec = c("time flies when you're having fun in 301",
            "time does not fly in 301, because it's not fun",
            "the Flyers are a hockey team in Philly")

grep(pattern = "fly", x = str.vec) 

grep("fly", str.vec, value = TRUE)

grep("Fly|flies", str.vec, value = TRUE)

# use grep to find the element in the character vector that matches "time"

grep("time", str.vec)
```

```{r}
str.vec.2 = c("time flies when you're having fun in 301",
              "fruit flies when you throw it",
              "a fruit fly is a beautiful creature",
              "how do you spell fruitfly?")

# what is the difference between each of these below?

grep("(time|fruit)(fly|flies)", str.vec.2, value=TRUE)
grep("(time|fruit) (fly|flies)", str.vec.2, value=TRUE)
grep("(time|fruit)  (fly|flies)", str.vec.2, value=TRUE)
```

**Metacharacters are special characters that have a special meaning, and are not interpreted literally.**

- Important example: square braces, used to indicate that we want to match anything inside the square braces, **for one character position.** 
- "[abcde]" matches the "a" in "Shawn"
    - "[369]" matches the "3" in "STT 315"
    - "[aeiou]" tries to match any lower case vowel
- A dash inside square braces is used to denote a range. 
    - "[a-e]" is the same as "[abcde]"
    - "[0-9]" is the same as "[0123456789]"
- Rules for combining regexes apply as before. 
    - "(Shawn|Santo) [A-J]512" tries to match "Shawn" or "Santo", then a space, then any upper case letter between A and J, then the number 512

**A summary of the R documentation on regex. See `?regex`.**    

- "[[:alnum:]]" tries to match any alphanumeric character (same as "[a-zA-Z0-9]")
- "[[:punct:]]" tries to match any punctuation mark
- "[[:space:]]" tries to match any white space character (including tabs and line breaks)
- A caret inside square braces negates what follows. 
    - "[^0-9]" tries to match anything but a number between 0 and 9
    - "[^aeiou]" tries to match anything but a lower case vowel
- A period "." tries to match any character (don't even need square braces)


Think about what is happening in each of the below uses of the `grep` function. Note the regular expressions that are used.


```{r}
str.vec.3 = c("C409WH","c409WH","Wells.Hall","C409")

grep("[A-Z][0-9]", str.vec.3, value=TRUE)

grep("[a-z][0-9][0-9][0-9][A-Z]", str.vec.3, value=TRUE)

grep("[^0-9]", str.vec.3, value=TRUE)

grep("[^0-9][^0-9]", str.vec.3, value=TRUE)

grep("[^0-9][a-z]", str.vec.3, value=TRUE)

grep("[A-Z][^0-9][^0-9][0-9]", str.vec.3, value=TRUE)

grep("[:punct:]", str.vec.3, value = TRUE)

grep("[[:punct:]]", str.vec.3, value = TRUE)
```

Use regular expressions to split `big.cats` at
  
  1. each "e".
  2. each vowel.
  
```{r}
strsplit(big.cats, split = "[e]")

strsplit(big.cats, split = "[aeiou]")
```

Use regular expressions to split `str.vec3` at

  1. each number.
  2. each lower case letter.
  3. each period.
  4. two consecutive numbers.
  5. where a capital letter immediately preceeds a digit
  

```{r}
strsplit(str.vec.3, split = "[0-9]")

strsplit(str.vec.3, split = "[a-z]")

strsplit(str.vec.3, split = "[.]")

strsplit(str.vec.3, split = "[0-9][0-9]")

strsplit(str.vec.3, split = "[A-Z][0-9]")

```


### A working example

Let's look at the all-time men's top 100m dash times. Ultimately, we want to extract the top 1000 times from this all-time list.

```{r}
sprint.times <- readLines("http://www.alltime-athletics.com/m_100ok.htm")
head(sprint.times)
sprint.times[81:88]
length(sprint.times)
```

Use `grep` to help find the first element in `sprint.times` that corresponds to "Usain Bolt". Save this as `sprint.start`.
```{r}
sprint.start <- min(grep("Usain Bolt", sprint.times))
```

Only keep the elements in `sprint.times` between `sprint.start` and 999 + `sprint.start`, inclusive. Save the object as `sprint.t1000`.

```{r}
sprint.t1000 <- sprint.times[sprint.start:(999+sprint.start)]
length(sprint.t1000)

```

Now, to extract the times we will use the `regmatches` function in R along with a regex time pattern that we create. Once you have done the above parts you can remove the chunk option of `eval = FALSE`.

`regexpr` returns an integer vector of the same length as `sprint.t1000` giving the starting position of the first match.

```{r }
time.pattern = "[[:space:]]+(9|(10))\\.[0-9]{2}([[:space:]]|A)"

match.data <- regexpr(time.pattern, sprint.t1000)

sprint.t1000 <-  regmatches(sprint.t1000, m = match.data)

length(sprint.t1000) # length should be 1000

head(sprint.t1000)
```

We need to clean this up a little more. Look at element 135 and you will see an A appended to the time. This is to signify wind assistance. If you use `as.numeric` that value will result in NA. Let's try one more application of `regmatches`. Again, remove the chunk option `eval = FALSE` once you have done the above parts.

```{r }
sprint.t1000 = as.numeric(regmatches(sprint.t1000, regexpr("(9|(10))\\.[0-9]{2}", sprint.t1000)))

length(sprint.t1000) # length should be 1000

summary(sprint.t1000)
```

Finally, we have cleaned up the data. Use the `sprint.t1000` to create a plot in ggplot of time versus rank.

```{r}
library(ggplot2)

top.times <- data.frame(rank=c(1:1000), time=sprint.t1000)

ggplot(data = top.times, mapping = aes(x = rank, y = time, colour = time)) + geom_point()
```


