---
title: "STT 301: Data frames, missing data, and floating point arithmetic"
author: "Shawn Santo"
date: "September 17, 2018"
output: html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

## Introduction

---

**Learning objectives:**

  + Data frame subsetting
  + Data frame manipulation
  + Floating point arithmetic

---

**Required data: `gapminder.Rdata`**

---

Today we will investigate a data set from [http://www.gapminder.org](http://www.gapminder.org), 
a site which contains a wealth of data and visualizations related to 
health, wealth, population, etc. of countries of the world. The data set 
has already been created in a `.Rdata` file available on D2L.

```{r, cache = TRUE}
load("gapminder.Rdata")
ls() # see the variables in the environment
```


## Data frames

The only object in `gapminder.Rdata` is the data frame object `gapminder`.
The `str` function reports on the structure of an object in `R`. 
It's often useful to use `str` when working with a new dataset.

It is never okay to display large data sets in your .Rmd file. No one wants
to wait minutes for the HTML file to generate when the .Rmd file is knit.

```{r}
str(gapminder)
```

We see that the `gapminder` data frame has six variables: `country` (a factor), 
`year` (an integer), `pop` (numeric), `continent` (a factor), 
`lifeExp` (numeric), and `gdpPercap` (numeric). 

Extract the third row of the data frame?

```{r}
gapminder[3, ]
```

Extract the first 50 elements of the variable `year`?

```{r}
gapminder$year[1:50]
```

There are a lot of repeated values in `year`. The `unique` function reports 
on the unique values in a variable. Display the unique values for the 
variable `year`.

```{r}
unique(gapminder$year)
```

Next, we will work with two particular years, 1952 and 2002, from the dataset. 
We will use the `subset` function to extract data frames containing data 
from these two years. Below is how to extract the data for 1952. Recall that
`==` is used for checking equality. If you need to remind yourself of the
subset function type `?subset` in the console.

```{r}
gapminder1952 <- subset(gapminder, year == 1952)
str(gapminder1952)
```

Now extract the data for 2002, and assign the resulting data frame to an 
object named `gapminder2002`.

```{r}
gapminder2002 <- subset(gapminder, year == 2002)
str(gapminder2002)
```

Let's create comparison boxplots of life expectancy for the five continents 
in the dataset for 1952.

```{r}
boxplot(lifeExp ~ continent, data = gapminder1952)
```

What do you observe from these boxplots?

**In both plots Oceania has the highest median life expectancy and Africa**
**has the lowest median life expectancy.**

Create similar boxplots for 2002. 

```{r}
boxplot(lifeExp ~ continent, data = gapminder2002, ylab = "Life Expectancy")
```

Do any obvious differences show up between the two years? 

**The median life expectancies for Americas and Asia increased dramatically**
**from 1952 to 2002.**

Next, calculate the median life expectancy, for both 1952 and 2002. 

```{r}
median(gapminder1952$lifeExp)
median(gapminder2002$lifeExp)
```

What do you notice?

**The median life expectancy increased by about 1.5 times from 1952 to 2002.**

Next, calculate the mean and median [GDP](https://en.wikipedia.org/wiki/Gross_domestic_product) 
for both years. 
```{r}
c(mean(gapminder1952$gdpPercap), median(gapminder1952$gdpPercap))

c(mean(gapminder2002$gdpPercap), median(gapminder2002$gdpPercap))
```

Again, what do you notice?

**Both have increased by about 2.7 times over the 50 years since 2002.**

Choose a country of interest. Create a data frame which only contains data from
that country, and then draw scatter plots of the life expectancy, and of the 
GDP, both versus the year. Recall that you can use the `plot` function to create
a scatter plot in `R`. Try and tidy your plots by using some of the available
graphical parameter options in the `plot` function.

```{r}
usa <- subset(gapminder, country == "United States")
usa

plot(usa$year, usa$lifeExp, xlab = "Year", ylab = "Life Expectancy")

plot(usa$year, usa$gdpPercap, xlab = "Year", ylab = "GDP per Capita")

```

## NA, NaN, Inf, and -Inf

For each of the following expressions, first guess what the output will be 
when the expression is evaluated in `R`, then enter the expression in `R`.

a. 1/0
b. 0/0
c. 1/0 + 5
e. 1/0 - 1/0
e. -2^(10000)
f. exp(Inf)
g. exp(-Inf)
h. 0^0
i. Inf^0
j. 0^Inf
k. log(0)
l. log(-Inf)

```{r}
1/0
0/0
1/0 + 5
1/0 - 1/0
-2^(10000)
exp(Inf)
exp(-Inf)
0^0
Inf^0
0^Inf
log(0)
log(-Inf)
```

## Floating point arithmetic

`R`, as does most software, uses floating point arithmetic, which is not the 
same as the arithmetic we learn. Computers cannot represent all 
numbers exactly. This has important consequences when we test for equality 
of numbers.

Below are a few examples.

First, would you expect `0.2 == 0.6/3` to return TRUE or FALSE? Test it out.

```{r}
0.2 == 0.6/3
```

Next consider the vector `point3` created below.

```{r}
point3 = c(0.3, 0.4 - 0.1, 0.5 - 0.2, 0.6 - 0.3, 0.7 - 0.4)
point3
```

Test whether each element of `point3` is equal to 0.3

```{r}
0.3 == point3
```

In calculus we learned that 
$$ e = \lim_{n\to\infty} (1 + 1/n)^n. $$

So in principle we should be able to get closer and closer approximations to 
$e$ by using larger and larger values of $n$ in the above expression. 
Let's experiment.

```{r}
n <-  100
(1 + 1/n)^n

n <-  10000
(1 + 1/n)^n

n <-  1000000
(1 + 1/n)^n

n <- 100000000
(1 + 1/n)^n
```

So far things are looking good. Investigate what happens as you increase $n$. 
Each time multiply $n$ by 10. Continue until you get to
10000000000000000. What happens?

```{r}
n <- 100000000
(1 + 1/n)^n

n <- 1000000000
(1 + 1/n)^n

n <- 10000000000
(1 + 1/n)^n

n <- 10000000000000000
(1 + 1/n)^n
```

**The answer eventually results in  1, because in `R` for extremely large `n`,**
**`1/n` results in 0.**