---
title: "STT 301: Data frames, missing data, and floating point arithmetic"
author: "Rellika Kisyula"
date: "September 17, 2018"
output: html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

## Introduction

---

**Learning objectives:**

  + Data frame subsetting
  + Data frame manipulation
  + Floating point arithmetic

---

**Required data: `gapminder.Rdata`**

---

Today we will investigate a data set from [http://www.gapminder.org](http://www.gapminder.org), 
a site which contains a wealth of data and visualizations related to 
health, wealth, population, etc. of countries of the world. The data set 
has already been created in a `.Rdata` file available on D2L.

```{r, cache = TRUE}
load("gapminder.Rdata")
ls() # see the variables in the environment
```


## Data frames

The only object in `gapminder.Rdata` is the data frame object `gapminder`.
The `str` function reports on the structure of an object in `R`. 
It's often useful to use `str` when working with a new dataset.

It is never okay to display large data sets in your .Rmd file. No one wants
to wait minutes for the HTML file to generate when the .Rmd file is knit.

```{r}
str(gapminder)
```

We see that the `gapminder` data frame has six variables: `country` (a factor), 
`year` (an integer), `pop` (numeric), `continent` (a factor), 
`lifeExp` (numeric), and `gdpPercap` (numeric). 

Extract the third row of the data frame?

```{r}
gapminder[3,]
```

Extract the first 50 elements of the variable `year`?

```{r}
head(gapminder$year, 50)
```

There are a lot of repeated values in `year`. The `unique` function reports 
on the unique values in a variable. Display the unique values for the 
variable `year`.

```{r}
unique(gapminder$year)
```

Next, we will work with two particular years, 1952 and 2002, from the dataset. 
We will use the `subset` function to extract data frames containing data 
from these two years. Below is how to extract the data for 1952. Recall that
`==` is used for checking equality. If you need to remind yourself of the
subset function type `?subset` in the console.

```{r}
gapminder1952 <- subset(gapminder, year == 1952)
str(gapminder1952)
```

Now extract the data for 2002, and assign the resulting data frame to an 
object named `gapminder2002`.

```{r}
gapminder2002 <- subset(gapminder, year == 1952)
str(gapminder2002)
```

Let's create comparison boxplots of life expectancy for the five continents 
in the dataset for 1952.

```{r}
boxplot(lifeExp ~ continent, data = gapminder1952, col="cyan")
```

What do you observe from these boxplots?
**Sol: The boxplots shows different life expectancy of different continents. Each boxplot is segmented into lower quantile, median, upper quantile.**
Create similar boxplots for 2002. 

```{r}
boxplot(lifeExp ~ continent, data = gapminder2002, col="yellow")
```

Do any obvious differences show up between the two years? 
**Sol: There is no obvious differences in these two years**
Next, calculate the median life expectancy, for both 1952 and 2002. 

```{r}
median_life_expectancy1952 =median(gapminder[gapminder$year == 1952,]$lifeExp)
cat("The median life expectancy, for 1952 :", median_life_expectancy1952, "\n")

median_life_expectancy2002 =median(gapminder[gapminder$year == 2002,]$lifeExp)
cat("The median life expectancy, for 2002 :", median_life_expectancy2002, "\n")
```

What do you notice?
**The median life expectancy of 2002 is 70.85 which is higher than 45.14 of 1952**

Next, calculate the mean and median [GDP](https://en.wikipedia.org/wiki/Gross_domestic_product) 
for both years. 
```{r}
median_gdp1952 =median(gapminder[gapminder$year == 1952,]$gdpPercap)
cat("The median GDP, for 1952 :", median_gdp1952, "\n")

median_gdp2002 =median(gapminder[gapminder$year == 2002,]$gdpPercap)
cat("The median GDP, for 2002 :", median_gdp2002, "\n")

# mean of the GDP

mean_gdp1952 =mean(gapminder[gapminder$year == 1952,]$gdpPercap)
cat("The mean GDP, for 1952 :", mean_gdp1952, "\n")

mean_gdp2002 =mean(gapminder[gapminder$year == 2002,]$gdpPercap)
cat("The mean GDP, for 2002 :", mean_gdp2002, "\n")
```

Again, what do you notice?
**Both the median GDP and mean of GDP for 2002 is (9917.84, 5319.81) are higher than (1968.53, 3725.28 ) of 1952**

Choose a country of interest. Create a data frame which only contains data from
that country, and then draw scatter plots of the life expectancy, and of the 
GDP, both versus the year. Recall that you can use the `plot` function to create
a scatter plot in `R`. Try and tidy your plots by using some of the available
graphical parameter options in the `plot` function.

```{r}
kenya <- gapminder[gapminder$country == "Kenya",]
plot(kenya$year, kenya$gdpPercap, type = "p", main = "GPD per Capita in Kenya", xlab = "Years", ylab = "GDP per Capita", s=8, col="red")

plot(kenya$year, kenya$lifeExp, type = "p", main = "Life Expectancy in Kenya", xlab = "Years", ylab = "Life Expectancy", s=8, col="red")
```

## NA, NaN, Inf, and -Inf

For each of the following expressions, first guess what the output will be 
when the expression is evaluated in `R`, then enter the expression in `R`.

a. 1/0
b. 0/0
c. 1/0 + 5
e. 1/0 - 1/0
e. -2^(10000)
f. exp(Inf)
g. exp(-Inf)
h. 0^0
i. Inf^0
j. 0^Inf
k. log(0)
l. log(-Inf)

```{r}
1/0 # Inf
0/0 # NaN
1/0 + 5 # inf
1/0 - 1/0 # Nan
-2^(10000) # -Inf
exp(Inf) # Inf
(-Inf) # -Inf
0^0 # 1
Inf^0 # 1
0^Inf # 0
log(0) # -Inf
log(-Inf) # NaN
```

## Floating point arithmetic

`R`, as does most software, uses floating point arithmetic, which is not the 
same as the arithmetic we learn. Computers cannot represent all 
numbers exactly. This has important consequences when we test for equality 
of numbers.

Below are a few examples.

First, would you expect `0.2 == 0.6/3` to return TRUE or FALSE? Test it out.
`False`
```{r}
0.2 == 0.6/3 # FALSE
```

Next consider the vector `point3` created below.

```{r}
point3 = c(0.3, 0.4 - 0.1, 0.5 - 0.2, 0.6 - 0.3, 0.7 - 0.4)
point3
```

Test whether each element of `point3` is equal to 0.3

```{r}
for (num in point3) {
  if (num ==0.3){
    print("TRUE")
  }else {
    print ("FALSE")
  }
}
```

In calculus we learned that 
$$ e = \lim_{n\to\infty} (1 + 1/n)^n. $$

So in principle we should be able to get closer and closer approximations to 
$e$ by using larger and larger values of $n$ in the above expression. 
Let's experiment.

```{r}
n <-  100
(1 + 1/n)^n

n <-  10000
(1 + 1/n)^n

n <-  1000000
(1 + 1/n)^n

n <- 100000000
(1 + 1/n)^n
```

So far things are looking good. Investigate what happens as you increase $n$. 
Each time multiply $n$ by 10. Continue until you get to
10000000000000000. What happens?

```{r}
n <-  100
(1 + 1/n)^n

n <-  10000
(1 + 1/n)^n

n <-  1000000
(1 + 1/n)^n

n <- 100000000
(1 + 1/n)^n

n <- 1000000000
(1 + 1/n)^n

n <- 10000000000
(1 + 1/n)^n

n <- 100000000000
(1 + 1/n)^n

n <- 1000000000000
(1 + 1/n)^n

n <- 10000000000000
(1 + 1/n)^n

n <- 100000000000000
(1 + 1/n)^n

n <- 1000000000000000
(1 + 1/n)^n

n <- 10000000000000000
(1 + 1/n)^n
```

