---
title: "STT 301: Text Data"
author: "Shawn Santo"
date: "November 14, 2018"
output: html_document
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, comment = NA)
```

## Introduction

Learning Objectives

  - String/character review
  - Reading in text data
  - Working with text data
  
*************************

## Some string/text basics

#### Whitespaces

Whitespaces count as characters and can be included in strings:

- `" "` for space
- `"\n"` for newline
- `"\t"` for tab

```{r}
message <- "Dear Shawn,\n\n Text data is a little boring. \n\n Sincerely, Everyone"
```

#### Printing strings

To print to the console, use the `cat` function.

```{r}
message
cat(message)
```

#### Vectors of strings

Recall, that character is a basic data type in R (like numeric, or logical), so we can make vectors of out them.

```{r}
str.1 <- "statistical"; str.2 <- "computing"; str.3 <- "isn't that bad"
str.vec <- c(str.1, str.2, str.3) # Collect 3 strings
str.vec # All elements of the vector
str.vec[3] # The 3rd element
str.vec[-(1:2)] # All but the 1st and 2nd
head(str.vec, 2) # The first 2 elements
tail(str.vec, 2) # The last 2 elements
rev(str.vec) # Reverse the order
```

#### Matrices of strings

The same idea applies for matrices.

```{r}
str.mat <- matrix("", 2, 3) # Build an empty 2 x 3 matrix
str.mat[1,] <- str.vec # Fill the 1st row with str.vec
str.mat[2,1:2] <- str.vec[1:2] # Fill the 2nd row, only entries 1 and 2, with those of str.vec
str.mat[2,3] <- "isn't a fad" # Fill the 2nd row, 3rd entry, with a new string
str.mat # All elements of the matrix
t(str.mat) # Transpose of the matrix
```

#### Converting strings

Recall, we can make values in R into strings using `as.character`. What if we want to convert a string? It is not always easy or straightforward.

```{r}
as.numeric("0.5")
as.numeric("0.5 ")
as.numeric("0.5e-10")
as.numeric("Hi!")
as.logical("True")
as.logical("TRU")
as.numeric(c("0.5", "TRUE"))
```


Additionally, we can convert lower and uppercase strings by using the functions: `tolower` or `toupper`.


```{r}
tolower("HELLO")
toupper("no thank you")
toupper("WhaT WiLl HaPpEn HeRe")
```


*************************

## Reading in text data

#### Using `readLines`

Use `readLines` to read in the words from the Enron email data set. The Enron email data set is a massive text data set that has every email sent between managers prior to its exposure of accounting and corporate fraud. Save this as an object named `enron`. The data is available at: "https://archive.ics.uci.edu/ml/machine-learning-databases/bag-of-words/vocab.enron.txt"

```{r}
enron <- readLines("https://archive.ics.uci.edu/ml/machine-learning-databases/bag-of-words/vocab.enron.txt")

length(enron)

which(enron == "skilling")

enron[1:10]
```


#### Using `scan`

Next we will read in the novel *Sense and Sensibility* by Jane Austen using the `scan` function. We specify that we are reading in character data, and that newline characters are the separators.

```{r}
SandS <- scan("http://www.stt.msu.edu/~melfi/STT301/data/austen.txt", what = "character", sep = "\n")
```

```{r}
which(SandS == "CHAPTER 1")

which(SandS == "THE END")

SandS <- SandS[which(SandS == "CHAPTER 1"):which(SandS == "THE END")]
head(SandS)
tail(SandS)

SandS.str <- paste(SandS, collapse = " ")
length(SandS.str)

str(SandS.str)

```

*************************

## The `paste` function

Below is function to test if a given number is prime. Examples are provided below the function.

```{r}
is.prime <- function(num) {
   if (num == 2) {
      TRUE
   } else if (any(num %% 2:(sqrt(num)) == 0)) {
      FALSE
   } else { 
      TRUE
   }
}
```

```{r}
is.prime(2)
is.prime(4)
is.prime(17)
```

**Write a function called `print.primes` that takes one argument, `n`, and prints out all the prime numbers less than or equal to `n`. You will want to use the function `is.prime` given above and the `print` function.**

The ouput should be what you see below.

```
> print_primes(10)
[1] 2
[1] 3
[1] 5
[1] 7
```

```{r}
print.primes <- function(n){
  
  for (i in 1:n){
    if (is.prime(i)){
      print(i)
    }
  }
  
}

print.primes(10)
```


**Rewrite `print.primes` to produce output such as you see below. You will want to now make use of the `paste` function. If you are getting quotes and the [1] try using `cat` instead of `print`.**

```
Prime Number 1 is 2 
Prime Number 2 is 3 
Prime Number 3 is 5 
Prime Number 4 is 7
```

```{r}
print.primes2 <- function(n){
  j <- 1
  for (i in 1:n){
    if (is.prime(i)){
      cat(paste("Prime Number", j, "is", i, collapse = ","), "\n")
      j <- j + 1
    }
  }
  
}

print.primes2(10)
```



