---
title: "STT 301 Homework Assignment 2 Solutions"
author: "Shawn Santo"
date: "September 26, 2018"
output:
  html_document: default
---

```{r setup, include=F}
library(knitr)
opts_chunk$set(comment = NA, message = FALSE, warning = FALSE)
```

## Introduction

<em>The opioid epidemic is the deadliest drug overdose crisis in US 
history - on track to kill more people over the next ten years than
currently live in entire cities like Baltimore or Miami. More people
died last year from drug overdoses than American soldiers during the
Vietnam War. In recent times, the opioid crisis has become an issue that
the country can no longer ignore.

So far in 2018, anti-opioid ads on television have aired in congressional and gubernatorial races more than 50,000 times across 25 states, said The Wall
Street Journal. At this point in 2014, Kentucky was the only 
state with political ads mentioning opioids that 
aired only 70 times. In other words, political ads against 
opioids have skyrocketed roughly 714 times when compared to this time in 
the 2014 midterm election cycle.</em>

The above is an excerpt from an article, "Opioid Crisis Emerges As 
Top Campaign Theme For Midterms". 

Connecticut is one of the states with a 
high number of opiod deaths. For this assignment you will work with
Connecticut's open data on Accidental Drug Related Deaths 2012-2017. The website
that hosts the data and other related resources is available
[here](https://data.ct.gov/Health-and-Human-Services/Accidental-Drug-Related-Deaths-2012-2017/rybz-nyjw). You will need to read the website first to understand
the data and variables. There are some other interesting features on the 
website for data visualization. By the end of this course you will have the
capabilities to replicate and expand on those visualizations.

You will need the following two packages for a plot to generate at the end
of this assignment. Install the packages with `install.packages("dplyr")` and
`install.packages("ggplot2")`. Do this in your console and do not include it
in your .Rmd file. After installation you will want to load the packages with
the `library` function. Include the below two lines of code as your second
R chunk in your .Rmd file. Remember, a package only needs to be installed
once, but each time you want to use it you need to load it into your workspace.

```{r}
library(dplyr)
library(ggplot2)
library(maps)
```

Below is code to read in the data. The `opiod` object is a data frame with
4083 rows and 32 columns. This should be your third R chunk.

```{r cache=TRUE}
opiod <- read.csv("https://data.ct.gov/api/views/rybz-nyjw/rows.csv?accessType=DOWNLOAD",
                  stringsAsFactors = FALSE)
names(opiod)
```

***********************

### Question 1 - Data exploration and cleaning (6 points)

#### Part a (0.75 points)

  1. Create a data frame called `opiod_dem` that contains all the rows
     of `opiod`, but only contains the columns `Sex`, `Race`, `Age`.
  2. Compute the mean, median, and standard deviation for `Age` 
     from `opiod_dem`.
  3. Create a boxplot for the variable `Age` from `opiod_dem`. Include a label
     for the y-axis that reads, "Age at time of death".
  
```{r include=T}
opiod_dem <- subset(opiod, select = Sex:Age)

c(mean(opiod_dem$Age, na.rm = TRUE), 
  median(opiod_dem$Age, na.rm = TRUE),
  sd(opiod_dem$Age, na.rm = TRUE))

boxplot(opiod_dem$Age, ylab = "Age at time of death")
```

  
#### Part b (4 points)

The function `table` builds a contingency table of the counts at each 
combination of factor levels. Try it out with `table(opiod_dem$Sex)`. You
should notice that for 4 of the deaths a value was not provided in the data
for `Sex`. In the data set empty values are as "".

  1. Subset `opiod_dem` to remove the rows where the variable `Sex` has ""
     as a value. Save the resulting data frame as `opiod_dem_filter`.
  2. Subset `opiod_dem_filter` to remove the rows where the variable 
     `Race` has "" as a value. Save the resulting data 
     frame as `opiod_dem_filter`.
  3. Use the function `table` to build a table for the variable `Sex`
     from `opiod_dem_filter`.
  4. Use the function `table` to create a two way contingency table for the
     variables `Race` and `Sex` from `opiod_dem_filter`. Save the resulting
     table as an object named `dem`.
  5. Turn `dem` into a data frame with the following code:
     `dem_df <- data.frame(female = dem[, 1], male = dem[, 2])`.
  6. Add a column named `sums` to `dem_df` that is the row sum.
  7. From 6, calculate which race has the highest percentage of female deaths, 
     and which race has the highest percentage of male deaths.
  8. Repeat the calculation in 6, but only for races with a minimum of 
     100 deaths.
     
```{r include=T}
# parts 1 and 2
opiod_dem_filter <- subset(opiod_dem, subset = (Sex != "") & (Sex != " "))
opiod_dem_filter <- subset(opiod_dem_filter, subset = (Race != ""))

# parts 3 and 4
table(opiod_dem_filter$Sex)
dem <- table(opiod_dem_filter$Race, opiod_dem_filter$Sex)

# parts 5 and 6
dem_df <- data.frame(female = dem[, 1], male = dem[, 2])
dem_df$sums <- dem_df$female + dem_df$male

# part 7
rownames(dem_df[which.max(dem_df$female / dem_df$sums), ])
rownames(dem_df[which.max(dem_df$male / dem_df$sums), ])

# part 8
dem_df100 <- subset(dem_df, subset = sums >= 100)
rownames(dem_df100[which.max(dem_df100$female / dem_df100$sums), ])
rownames(dem_df100[which.max(dem_df100$male / dem_df100$sums), ])
```

#### Part c (1.25 points)

You will now get back to working with the data frame object `opiod`.

  1. Display the unique counties for the variable `Death.County`. You should
     observe that three counties do not make sense. One is "", another is
     "USA", and another is " FAIRFIELD" as opposed to "FAIRFIELD",
     note the extra space. You will correct each of these issues in 2-4.
  2. Update `opiod` so " FAIRFIELD" is "FAIRFIELD" for the 
     variable `Death.County`.
  3. Update `opiod` so "" is "NOT RECORDED" for the variable `Death.County`.
  4. Update `opiod` so "USA" is "NOT RECORDED" for the variable `Death.County`.
  5. Create a barplot with the function `barplot` for the number of deaths
     from the variable `Death.County`. As a hint, use the `table` function
     to generate counts for each county in `Death.County`. If you did steps 
     2-4 correctly, you should have nine categories. Include the graphical
     parameter `cex.names = 0.5` to shrink the county names so they fit in
     the plot window.
  6. Update `opiod` so "" values for the variable `Race` are "Unknown".
  
  
```{r include=T}
# part 1
unique(opiod$Death.County)

# parts 2 - 4
opiod$Death.County[opiod$Death.County == " FAIRFIELD"] <- "FAIRFIELD"
opiod$Death.County[opiod$Death.County == ""] <- "NOT RECORDED"
opiod$Death.County[opiod$Death.County == "USA"] <- "NOT RECORDED"

# part 5
barplot(table(opiod$Death.County),
        xlab = "County", ylab = "Death count", main = "Death counts by county",
        cex.names = 0.4)

#part 6
opiod$Race[opiod$Race == ""] <- "Unknown"
```
  

### Question 2 - County summary function (3 points)

Create a function named `county.summary` that has two arguments: `df` and
`county`. The `opiod` data frame will be passed to the
argument `df`, and a county in Connecticut will be passed to the `county`
argument as a character. See below for some examples of `county.summary` in 
action.

Your function, `county.summary`, should return a list with the following 
information:

  + total deaths in the specified county,
  + mean age of death in the specified county,
  + median age of death in the specified county,
  + a demographics data frame for county deaths based on race and gender.
  
An idea of how the function should work: the county name passed into the
function should filter the data frame based on the variable 
`Death.County`; compute the necessary information from the filtered 
data frame; return the results in a list and in the format you see below. 
Include a check at the begining of your function that will give an error to the
user if a county is entered that is not in Connecticut.


```{r echo=T}
county.summary <- function(df, county){
  # check if county is in CT
  if (!(county %in% unique(df$Death.County))) {
    stop("County is not located in CT!")
  }
  # create a county data frame
  df.county <- subset(df, subset = (Death.County == county))
  
  # compute summary statistics
  total.deaths <- dim(df.county)[1]
  mean.age <- mean(df.county$Age, na.rm = TRUE)
  median.age <- median(df.county$Age, na.rm = TRUE)
  demograph <- table(df.county$Race, df.county$Sex)
  
  if (dim(demograph)[2] == 3) {
    colnames(demograph) <- c("Unknown", "Female", "Male")
  }

  # aggregate results in a list
  county.results <- list(total.deaths = total.deaths,
                         mean.age = mean.age,
                         median.age = median.age,
                         demographics = demograph)
  return(county.results)
}
```

```{r error=TRUE}
county.summary(df = opiod, county = "LITCHFIELD")

county.summary(opiod, "HARTFORD")

county.summary(opiod, "hartford")

county.summary(opiod, "INGHAM")
```

### Question 3 Choropleth map (1 point)

#### Part a

  1. Use the `opiod` data frame to create a data frame the contains two columns.
     Column 2 should contain the counts of total deaths in that county. 
     Column 1 should contain the county's name. You should work with the
     `Death.County` variable.
  2. Add the names "subregion" and "count" to the data frame in 1.
  3. Remove the "NOT RECORDED" row from your data frame.
  4. Pass your data frame into the function `ct.choropleth` given below.
     Explain ways to improve the plot to make it more informative.
  
```{r include=T}
# parts 1 - 3
county.df <- data.frame(table(opiod$Death.County))
names(county.df) <- c("subregion", "count")
county.df <- county.df[county.df$subregion != "NOT RECORDED", ]
```

```{r}

ct.choropleth <- function(df){

  # generate county and state boundaries
  ct.state <- map_data("state", region = "connecticut")
  ct.county.df <- map_data("county", region = "connecticut")
  
  # convert county names to lower case
  county.df <- mutate_all(df, funs(tolower))
  
  # merge data frames to pass a single data frame to ggplot
  choropleth <- inner_join(ct.county.df, county.df, by = "subregion")
  
  # convert counts to type numeric
  choropleth$count <- as.numeric(choropleth$count)
  
  # generate choropleth
  ct.plot <- ggplot(choropleth, aes(long, lat, group = group)) + 
    geom_polygon(aes(fill = count), alpha = 0.75, color = "white") + 
    geom_polygon(data = ct.county.df, colour = "white", fill = NA) +
    geom_polygon(data = ct.state, color = "black", fill = NA)+
    scale_fill_gradient2(low = "yellow", mid = "orange", high = "red") +
    ggtitle("Opiod deaths in Connecticut by county") +
    labs(fill = "Deaths") +
    theme_void()

  return(ct.plot)
}

ct.choropleth(county.df)

```

  1. Add county names
  2. Interactive plot that shows county statistics when a county is clicked on
  3. Add cities
  4. Add death and residence information
  5. Make 5 plots, 1 for each year, to see the changes across time