---
title: "STT 301: Lists, logical subsetting, and patterned data"
author: "Shawn Santo"
date: "September 19, 2018"
output: html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

## Introduction

---

**Learning objectives:**

  + Manipulating lists
  + Logical subsetting
  + Data frame manipulation
  + Plot options

---

**Required data: `gapminder.Rdata`**

---

## Lists

Today we will pull our first data set from a website and read it into `R`
using the built-in function `read.table`.

```{r}
lsd <- read.table("http://www.stat.ufl.edu/~winner/data/lsd.dat", header = F)
lsd
```

This small data set contains math test scores with tissue concentration of 
lysergic acid diethylamide (LSD). The data set is from Wagner JG, e. 
Correlation of performance test scores with "tissue concentration" of lysergic 
acid diethylamide in human subjects. - PubMed - NCBI .
Ncbi.nlm.nih.gov. First, we will assign our data column names, 
look at the structure, and plot the data.

```{r}
colnames(lsd) <- c("concentration", "mathscore")
str(lsd)

plot(lsd$concentration, lsd$mathscore, 
     xlab="LSD Concentration", ylab="Math Score", 
     col="blue", pch=16)

# pch is a graphical parameter that can have value 0 to 25 - it will change 
# the points to a different shape/object

# col is a graphical parameter that will set the color of the points
```

Compute the correlation between the two quantitative variables.

```{r}
cor(lsd)
```

Based on the scatter plot it does not appear that any of the linear model
assumptions are violated. The `lm` function fits a linear model in `R` 
and returns a list containing elements such as the coefficients, residuals, etc.

Next we fit a linear model, assign it to the `R` object `lsd_lm`, 
ask whether the object is a list, and obtain the names of the 
elements of `lsd_lm`.

We will learn about linear models and how to use the `lm` function in detail
later in the course. For now our focus is on the output and working with lists.

```{r}
lsd_lm <- lm(mathscore ~ concentration, data = lsd)
is.list(lsd_lm)
names(lsd_lm)
```

We see that `lsd_lm` is a list with 12 components. 

We can extract the components of `lsd_lm` in several ways. First, we will use
the `$` operator. For example, `lsd_lm$residuals` returns the residuals 
from the model. Use this method to extract the coefficients and the residuals
from the list `lsd_lm`. Assign the results to `R` objects called 
`lsd_lm_coef1` and `lsd_lm_resid1`, respectively.

```{r}
lsd_lm_coef1 <- lsd_lm$coefficients
lsd_lm_resid1 <- lsd_lm$residuals
```

Now display the coefficients and the residuals. 
```{r}
lsd_lm_coef1
lsd_lm_resid1
```

We might expect that the resulting objects are vectors. 
Use `is.vector` to test whether this is the case.

```{r}
c(is.vector(lsd_lm_coef1), is.vector(lsd_lm_resid1))
```

There are several other ways to access parts of a list. 
Below are a few ways to extract the coefficients.

1. `lsd_lm$coefficients`
2. `lsd_lm[["coefficients"]]`
3. `lsd_lm["coefficients"]`

We could also use the number `1` in place of `"coefficients"` in (2) and (3).

We've already used method 1 and assigned the result to `lsd_lm_coef1`. 
Use the second and third methods, and assign the results to 
`lsd_lm_coef2` and `lsd_lm_coef3`, respectively.
Also display all three results below.

```{r}
lsd_lm_coef2 <- lsd_lm[["coefficients"]]
lsd_lm_coef3 <- lsd_lm["coefficients"]

lsd_lm_coef1
lsd_lm_coef2
lsd_lm_coef3
```

Do the resulting objects look different? Apply `is.vector` and `is.list` on each
of the three objects.

```{r}
c(is.vector(lsd_lm_coef1), is.vector(lsd_lm_coef2), is.vector(lsd_lm_coef3))

c(is.list(lsd_lm_coef1), is.list(lsd_lm_coef2), is.list(lsd_lm_coef3))
```

You should have discovered that methods 1 and 2 returned a vector, but method 3
returned a list. The difference may seem subtle and not particularly important,
but it will become much more important when we start to write our own functions.

To see a summary of the contents associated with the `lm` object we can use 
the `summary()` function.
```{r}
summary(lsd_lm)
```

We will now add the least squares regression line to the plot that we
previously created.
```{r}
plot(lsd$concentration, lsd$mathscore, 
     xlab="LSD Concentration", ylab="Math Score", 
     col="purple", pch=17)

abline(lsd_lm)
```


## Subsetting with logical vectors

Subsetting with logical vectors is an essential skill. We'll practice this 
using the `gapminder` data frame. First we use the `load()` function to 
load the .Rdata file. You downloaded this file last class, but if you need it
again it is available on D2L.

```{r}
load("gapminder.Rdata")
str(gapminder)
```

Below are two examples that use logical subsetting.

First, we compute the mean population of France for the years contained 
in the data set.

```{r}
mean(gapminder$pop[gapminder$country == "France"])
```

Next, we extract the countries in the continent `Africa`. Since the countries
are repeated for each of the  years, we'll use `unique` to list the
countries only once.

```{r}
unique(gapminder$country[gapminder$continent == "Africa"])
```

We can combine conditions with the operators `&` and `|` which represent "and"
and "or". For example, here's how to extract all the countries whose population
was more than 150 million in the year 1992.

```{r}
gapminder$country[(gapminder$pop > 150000000) & (gapminder$year == 1992)]
```

Use these ideas to extract the population of all countries whose 
life expectancy is more than 70 years for the year 1967.

```{r}
gapminder$pop[(gapminder$lifeExp>70) & (gapminder$year==1967)]
```

Another useful operator is `%in%`. 

```{r}
x <- 5:10
y <- c(3, 5, 6, 9, 12, 15)

x %in% y
y %in% x
```

For example, here's how to compute the mean
population for the United States, Mexico, and Canada during 1997.

```{r}
mean(gapminder$pop[gapminder$country %in% c("United States", "Mexico", "Canada") &
                     gapminder$year == 1997])
```

Subsetting also can be used to change values of existing `R` objects 
as in the following example.

```{r}
dd = data.frame(x = c("dog", "cat", "oink", "pig", "oink", "cat", "dog", "cat", "dog"), 
                y = c("dog", "cat", "cat", "pig", "cow", "dog", "dog", "cat", "pig"),
                stringsAsFactors = FALSE)
dd

dd$x[dd$x == "oink"] <- "pig"
dd

dim(dd)
dd$same <- rep("no", dim(dd)[1])
dd

dd$same[dd$x == dd$y] <- "yes"
dd
```

Next, we will add a variable called `G8` to the `gapminder` data frame, which
will be equal to 1 or 0 depending on whether the country is in the G8 group of
nations: France, Germany, Italy, the United Kingdom, Japan, the United States, 
Canada, and Russia.

First, create a variable called `G8` - all of whose values are 0. 
This variable should be added to the data frame and can be done
by creating the object `gapminder$G8`.

```{r}
gapminder$G8 <- rep(0, dim(gapminder)[1])
```

Now, set `G8` equal to 1 if the country is in the G8. 
One way to do this is to give eight separate `R` commands, one for each of the
G8 countries. Another method is to use `%in%` and complete the work in one line
of `R` code.

```{r}
gapminder$G8[gapminder$country 
             %in% c("France","Germany","Italy","United Kingdom",
                    "Japan","United States","Canada","Russia")] <- 1
```


## Patterned data

We'll plot the function $f(x) = \cos(x)$ for $x$ between $-2\pi$ and $2\pi$.

First, create a vector `x` containing 1000 equally spaced values between 
$-2\pi$ and $2\pi$ using the `seq` function. In `R` you can access $\pi$ by
typing `pi`.

```{r}
x <- seq(from=-2*pi, to=2*pi, length.out = 1000)
```

Next, use the `plot` function to plot $f(x)$ defined above. The first argument
to `plot` should be the vector `x` we just created. The second argument 
should be `cos(x)`. 

```{r}
plot(x, cos(x))
```

The graphic is okay, but since we're plotting the graph of a function we 
should connect the points rather than plotting each of the points. 
Add the argument `type = "l"`.

```{r}
plot(x, cos(x), type="l", col = "blue")
```

The `rep` function replicates its first argument. 
Important arguments include `length.out`, `times`, and `each`. This can be 
very useful for creating vectors with respect to some pattern. Run the below
chunk to investigate what happens as different arguments are specified.

```{r}
ltrs <-  c("a", "b", "q", "z")
rep(ltrs, length.out = 9)
rep(ltrs, each = 3)
rep(ltrs, times = 3)
rep(ltrs, times = c(3, 2, 4, 7))
```



