---
title: "FOR/STT 875, Exercise 3"
output: html_document
---

## Learning objectives
  + create, subset, and manipulate vector contents and attributes
  + summarize vector data using R `table` and other functions
  + generate basic graphics using vector data

## Overview
Modify this R Markdown file by filling in the code needed to answer the questions. The Exercise 3 video provides some guidance for completing this exercise.

## Submission instructions
Upload your exercise_3.Rmd and exercise_3.html files to the Exercise 3 D2L dropbox. 

## Grading
You will receive full credit if your R Markdown document: 1) compiles without error; and 2) correctly completes the `# TODO:` code chunks. Also, please, fill in the feedback [questions](#questions) at the end of the exercise.  

## Getting started
First we'll set code chunk options to be applied to the entire document (i.e., global options). These global options can be overridden by the options in an individual code chunk. Typically we would use `echo = FALSE` in this chunk, since we wouldn't want it seen in the final document, but here we display the chunk. The global knitr options change how the R code is displayed in the resulting html document. Specifically, `comment = NA` removes the `##` comment that come before printed code and `tidy = TRUE` causes the printed code to wrap neatly.

```{r global_options}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

## Z&uuml;richberg Forest inventory data set
The data come from the Z&uuml;richberg Forest which overlooks Lake Z&uuml;rich in Switzerland. These data comprise a complete enumeration of all trees in the Z&uuml;richberg Forest including species (`spp`), diameter at breast height (`dbh`; the tree's diameter in cm at 1.4 m from the ground), and volume (`vol`; the volume of the tree's stem in $m^3$).

We'll read in three vectors that hold tree species, diameter, and volume. The element ordering in the vectors correspond to the same tree (i.e., element 1 in each vector describes the first tree's characteristics).

Read in the Z&uuml;richberg Forest data.

```{r}
con <- url("http://blue.for.msu.edu/FOR875/data/ZF_trees.gz")
load(con)
close(con)
rm(con)
ls()
```

List the first 25 species in `spp`.

```{r}
# TODO: List the species here
head(spp, 25)
```

List the first 25 diameter measurements in `dbh`.

```{r}
# TODO: List the DBH here
head(dbh, 25)
```

List the first 25 volume measurements in `vol`.

```{r}
# TODO: List the volume here
vol[1:25]
```

The `table` function builds a contingency table. For example,

```{r}
animals <- c("dog", "cat", "dog", "cat", "horse", "dog")
table(animals)
```

Create a table that gives the number of times each species occurred in the data set.

```{r}
# TODO: Show the table here
table(spp)
```

By default the table is ordered alphabetically by the name of the species. It might be more useful to order by the number of times a species occurs. Apply the `sort` function to the table to achieve this, and save the resulting table as `spp_table`.

```{r}
# TODO: Show the sorted table here
spp_table = sort(table(spp), decreasing = FALSE)
spp_table

```

By observation, which three species are most abundant?
```{r}
# TODO: Write the three most abundant spp here (comment out your answer using the # symbol)
# 1. Spruce with 2006
# 2. Beech with 1625
# 3. Maple with 477
```

Subset `spp_table` using `[]`to create a table called `top_spp` that contains the three most abundant species. Write your subset code so that it will always return the three most abundant species, even if the number of species changes. Do not hard code `6:8` in the square brackets, such as `spp_table[6:8]`; rather, use the `length` function and some basic math in place of hard coded vector positions.

```{r}
# TODO: Create the top_spp table here
startSubset = length(spp_table)-2
endSubset = length(spp_table)
top_spp = spp_table[startSubset:endSubset]
top_spp
```

Draw a bar graph of the `top_spp` table using the `barplot` function.

```{r}
# TODO: Create the bar graph here
barplot(top_spp)
```

Now add a y-axis label "Number of trees" to the barplot.
```{r}
# TODO: Create the bar graph with y-axis labels here
barplot(top_spp, ylab ="Number of trees" )
```

Now make a table of species abundance, and sort in decreasing order, by giving the `decreasing = TRUE` argument in the `sort` function.

```{r}
# TODO: Show the sorted table here
treesAbundance = spp_table = sort(table(spp), decreasing = TRUE)
treesAbundance
```

Use an R function to figure out how many trees are in the Z&uuml;richberg Forest data.
```{r}
# TODO: Compute the answer here
# Get length of one of the vectors
length(spp)
```

The vectors listing tree diameter and volume would be more informative if the species were included. We can give names to each vector element using the `names` function. For example:
```{r}
animals
names(animals) <- c("Asta", "Felix", "Chopper", "Garfield", "Mr. Ed", "Spud")
animals
```

Use `names` to add the species as names to the `dbh` and `vol` vectors, and list the first ten elements of each vector with the names included.
```{r}
# TODO: Create the vectors with names and list the first ten elements of each here.
names(dbh) = spp
dbh[1:10]

names(vol) = spp
vol[1:10]
```

###More advanced subsetting

So far we have directly specified the elements of vectors that we want to extract, for example `spp[1]` or `dbh[length(dbh)]` or `spp[c(1,2,3,5)]`. More commonly we want to extract elements that meet a condition, such as all trees greater than some minimum DBH or all trees of a given species. For this we use subsetting with logical vectors, see Section 4.7 in the course book.

First some very simple examples using vectors of random numbers, and then you'll use these ideas for the tree vectors.

```{r}
set.seed(123) #Set the seed for reproducible random number generation
x <- rbinom(10, 20, 0.9) #Generate 10 binomial(20, 0.9) values
y <- rbinom(10, 20, 0.8) #Generate 10 binomial(20, 0.8) values
x
y
```

```{r}
x[x > 16] #Values of x which are larger than 16
x[x == 18] #Values of x which are equal to 18
length(x[x == 18]) #How many values of x are equal to 18?
x[y < 15] #Values of x for which the corresponding values of y are less than 15
x[x < y] #Values of x which are less than the corresponding values of y
x[x == 18] <- 0 #Replace 18 by 0 in x
x
```

Now we use logical subsetting to learn about the Z&uuml;richberg Forest data. 

First, use logical subsetting to count the number of maple trees in the forest.

```{r}
# TODO: Count the number of maple trees here
length(spp[spp=='maple'])
sum(spp=='maple')
```

Did you answer the above question using `sum(spp=="maple")`? If not, then figure out why the `sum` function returns the correct answer.

We can combine conditions with the operators `&` and `|` which represent "and" and "or" for vectors. Also `!` negates, so for example `!=` represents not equal to. Also `x %in% y` is very useful:

```{r}
letters
letters %in% c("a", "m", "q", "s")
```

#####Questions:

Use these ideas to: 

a. Get the DBH of larch and silver fir trees and assign them to `dbh_l_sf`. Use the `%in%` operator for subsetting.
b. Using your solution from part a, calculate the mean and standard deviation of the larch and sliver fir trees.
c. Repeat part a, but this time use the `|` operator (i.e., you should be able to get the same answer as part a).
d. What is the maximum DBH and volume among the beech trees?

```{r QuestionA}
# TODO: Answer to a here
dbh_l_sf = dbh[spp %in% c("larch", "silver fir")]

# Display the first 10 output to see if the answer is correct
dbh_l_sf[1:10]
```

```{r QuestionB}
# TODO: Answer to b here
length(dbh_l_sf)
mean(dbh_l_sf)
sd(dbh_l_sf)

# Answer: Mean of the larch and sliver fir trees is 38.83333
# Answer: Standard Deviation of the larch and sliver fir trees is 17.33933
```

```{r QuestionC}
# TODO: Answer to c here
dbh_l_sf2 =dbh[spp=='larch'|spp=='silver fir']
length(dbh_l_sf2)

# Display the first 10 output to see if the answer is correct
dbh_l_sf2[1:10]
```

```{r QuestionD}
# TODO: Answer to d here
maxDBH = max(dbh[spp=='beech'])
maxDBH

maxVOL = max(vol[spp=='beech'])
maxVOL

# Answer: Max Volume is 9.14
# Max DBH is 85
```

********

Congratulations! You've reached the end of Exercise 3.

###Questions?

If you have any lingering questions about the material in this document or in the corresponding chapter, put them here.

*Response: I think the wording of the last question is not as clear what we were supposed to do. If you look at Piazza, I think it is one of the most asked question for clarification. I can employ the concepts I learned but it is so hard to understand the wording,*

###Giv us your feedback!

1.  How do you feel you're doing in this class?
*Response: The class is engaging and not so hard to follow if you read the chapters well*

2.  What could be done to improve your learning experience?
*Response: I wish it was possible if you notice a particular question on Piazza has been asked quiet a lot regarding a particular problem, then probably epxlain the problem to us a bit more. Some students have the concepts but sometimes they arent really sure what the questions is asking so they cant really employ the concepts*

*********