---
title: "STT 301: More with K-nn"
author: "Shawn Santo"
date: "November 12, 2018"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, comment = NA)
```

## Introduction

Learning Objectives

  - K-nn
  - Linear regression
  
*************************

```{r}
library(tidyverse)
library(class)
library(caret)
```

## Wine Classification with K-nn

### Read in data

```{r}
wineurl <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data"

wine.df <- read.csv(file = wineurl, header = FALSE)

str(wine.df)

wine.df <- as_tibble(wine.df)

colnames(wine.df) <- c("origin", "alcohol", "acid", 
                       "ash", "alcalinity", "magnesium", 
                       "phenols", "flavanoids", "nonflavanoid",
                       "proanthocyanins","color.int", "hue",
                       "od", "proline")

wine.df$origin <- as.factor(wine.df$origin)
```

### Plot the data

Get an idea of which variables influence the wine's origin

```{r}
wine.df2 <- data.frame(origin = wine.df$origin, scale(wine.df[,-1]))
wine.df2 <- gather(wine.df2, type,  value, -1)
ggplot(data = wine.df2, mapping = aes(x = origin, y = value)) + 
  geom_boxplot() + facet_wrap(~as.factor(type))

ggplot(data = wine.df, mapping=aes(x = flavanoids, 
                                   y = phenols, 
                                   color =  origin, 
                                   shape = origin)) + 
  geom_point()

with(wine.df, table(origin))
```

### Training and testing data sets

Split the data into training and testing sets. Use `createDataPartition` from
the `caret` package. You will need to use the arguments `y`, `p`, and 
`list = FALSE`. Confirm that your training and testing data sets have 
approximately equivalent origin proportions between the sets. Set the seed
to 3033 if you want to obtain identical results.

```{r}
set.seed(3033)

train.sample <- createDataPartition(y = wine.df$origin, p = 0.7, list = FALSE)
training <- wine.df[train.sample, ]
test <- wine.df[-train.sample, ]

# training proportion
cat("Training data set proportions:", "\n")

with(training, table(origin) / dim(training)[1])

# testing
cat("Testing data set proportions:", "\n")

with(test, table(origin) / dim(test)[1])
```

### K-nn

Use the K-nn algorithm to predict the wine origin in the testing data set.
Feel free to use some or all of the wine features. Can you get a prediction
accuracy above 98%?

```{r}
# not scaled

knn.fit <- knn(train = training[-1],
               test = test[-1],
               cl = training$origin,
               k = 21)

table(knn.fit, test$origin)


# scaled
knn.fit <- knn(train = scale(training[-1]),
               test = scale(test[-1]),
               cl = training$origin,
               k = 21)

table(knn.fit, test$origin)

```

## Linear Regression

  1. Select a data set from [http://users.stat.ufl.edu/~winner/datasets.html](http://users.stat.ufl.edu/~winner/datasets.html) in the section "Linear Regression". You can read the data directly
  into R by following the code above for the `wine` data set.
  2. Fit a linear regression model with all variables. Which variables
     are significant? Refine the fit.
  3. Investigate the refined model fit. Look at $R^2$ and a plot of residuals
     compared to $y$.
  4. Make a prediction for $y$ given a set of $x$ values.
  
```{r}
# some data
hg <- read.csv("http://users.stat.ufl.edu/~winner/data/fishermen_mercury.csv",
               header = T)

dim(hg)

# some plots
plot(x = hg$fishmlwk, y = hg$TotHg)
plot(x = hg$weight, y = hg$TotHg)

# model fits
hg.fit <- lm(TotHg ~ fisherman + age + restime + height + 
               weight + fishmlwk + fishpart, data = hg)
summary(hg.fit)

hg.fit2 <- lm(TotHg ~ weight + fishmlwk, data = hg)
summary(hg.fit2)

plot(y = hg.fit2$residuals, x = hg$weight )


# a prediction
predict(object = hg.fit2, newdata = data.frame(weight = 70, fishmlwk = 1))

```

