---
title: "STT 301: More with K-nn"
author: "Rellika Kisyula"
date: "11/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Learning Objectives

* K-nn
* Linear regression

```{r}
packages = c("caret","tidyverse")
#if a package is installed, it will be loaded
#if any are not, the missing package(s) will be installed and loaded
package.check <- lapply(packages, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x, dependencies = TRUE)
        library(x, character.only = TRUE)
    }
})
#verify they are loaded
search()
```

## Wine Classification with K-nn

### Read in data
```{r}
wineurl <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine/wine.data"
wine.df <- read.csv(file = wineurl, header = FALSE)
str(wine.df)
```


```{r}
wine.df <- as_tibble(wine.df)

colnames(wine.df) <- c("origin", "alcohol", "acid", 
                       "ash", "alcalinity", "magnesium", 
                       "phenols", "flavanoids", "nonflavanoid",
                       "proanthocyanins","color.int", "hue",
                       "od", "proline")

wine.df$origin <- as.factor(wine.df$origin)
```

### Plot the data
Get an idea of which variables influence the wine’s origin
```{r}
wine.df2 <- data.frame(origin = wine.df$origin, scale(wine.df[,-1]))
wine.df2 <- gather(wine.df2, type,  value, -1)
ggplot(data = wine.df2, mapping = aes(x = origin, y = value)) + 
  geom_boxplot() + facet_wrap(~as.factor(type))

ggplot(data = wine.df, mapping=aes(x = flavanoids, 
                                   y = phenols, 
                                   color =  origin, 
                                   shape = origin)) + 
  geom_point()

with(wine.df, table(origin))
```


### Training and testing data sets
Split the data into training and testing sets. Use `createDataPartition` from the `caret` package. You will need to use the arguments `y`, `p`, and `list = FALSE`. Confirm that your training and testing data sets have approximately equivalent origin proportions between the sets. Set the `seed` to 3033 if you want to obtain identical results.

```{r}

```

