---
title: "STT 301 Homework Assignment 5"
author: "Shawn Santo"
date: "November 21, 2018"
output:
  html_document: default
  pdf_document: default
---
***********************

### Homework Assignment 5 is due Monday, December 3 at 12:40pm EST.

***********************

### Instructions

This assignment is to be done in groups using R Markdown. 
Groups are set as your project groups. 
One Rmd file per group should be submitted to the 
dropbox folder by the above deadline with each individual's name listed in the 
"author" section. 

Everyone in the group will earn the same grade except for free-riders.

### Rubric

- **Total**: 10 points.
- **Correctness**: Point values for the question and their respective parts 
  are listed. Partial credit is available. Hard-coded solutions will
  not receive full credit.
- **Knitting**: Deduction of 0.5 points if the Rmd file does not knit for 
  any reason.
- **Style**: Use a third-level header to off-set each question in your 
  solutions - as is done below. For questions with multiple parts (part a,
  part b, etc), use 
  fourth-level headers to off-set the parts in your solutions - as is done 
  below. Use code comments for subsubparts. 
  Coding style is very important. You will receive a deduction of up 
  to 1.0 point if you do not adhere to good coding style. What I am looking for
  in terms of style includes:
    - appropriate variable use and naming
    - appropriate function use
    - good code commenting
    - consistent code syntax
- **Code documentation**: Code should be well documented.
- **Late Submission**: Late homework will not be accepted.
    
*Please do not include the above Rubric, Instructions, and homework deadline*
*sections in your solutions.*

***********************


```{r echo=FALSE, global_options}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

### Question 1 (10 points)

The goal of this assignment is to develop a word cloud as in step 5 of this
[word cloud tutorial](http://www.sthda.com/english/wiki/text-mining-and-word-cloud-fundamentals-in-r-5-simple-steps-you-should-know). You do not need to follow all the steps
exactly because you will work with different data. Steps 4 and 5 are most salient.

The [Unabomber](https://en.wikipedia.org/wiki/Ted_Kaczynski) was identified, captured, and convicted in part through
linguistic analysis. His published manifesto was linked to earlier writings
provided by family members. Ted Kaczynski's full manifesto is available [here](https://www.josharcher.uk/static/files/2018/01/Industrial_Society_and_Its_Future-Ted_Kaczynski.txt). Use
the text analysis functions we learned to create a word cloud of the Unabomber's manifesto. Be sure to remove numbers and 
common words such as "the", "and", "a", "an", and so on. Also, put all
text in lowercase. Remember to document
your code well.


#### Part 1 - read in manifesto

```{r}

```

#### Part 2 - clean and organize manifesto

```{r}

```

#### Part 3 - create word cloud

```{r}

```


