---
title: "STT 301: More graphics with ggplot2"
author: "Shawn Santo"
date: "October 8, 2018"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='Figs/', warning=FALSE, message=FALSE, dev = "png", comment = NA)
```

## Introduction

---

**Learning objectives:**

  - `ggplot` function
    - `geom_bar`
    - `geom_histogram`
    - `geom_boxplot`
    - `geom_raster`
    - `geom_density`

---

The package `ggplot2` has many options and capabilities so you will probably find the following `ggplot2` help resources useful:

  1. [http://docs.ggplot2.org/current/](http://docs.ggplot2.org/current/)
  2. the graphics portion of the R cookbook site at http://www.cookbook-r.com/Graphs/
  3. http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/
  4. RStudio "Data Visualization with ggplot2" cheatsheet
  4. Google

**********

## Part 1
### Bar plots - Diamonds data

The ggplot2 package comes with a data set called `diamonds`. Let's look at 
it below. To obtain further details type `?diamonds` in your console window.

```{r}
library(ggplot2)
str(diamonds)
```

Last time you created scatter plots such as the one below by making use of `geom_point`.

```{r}
diamond.plot <- ggplot(data=diamonds, aes(x=carat, y=price, colour = color))
diamond.plot + geom_point() + facet_wrap(~cut)
```

Use `geom_bar` to create the bar plots you see below.

#### Plot a

```{r echo = T}
diamond.bp <- ggplot(data = diamonds, mapping = aes(x = cut))
diamond.bp + geom_bar()
```

#### Plot b

```{r echo = T}
diamond.bp2 <- ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity))
diamond.bp2 + geom_bar()
```

#### Plot c

Details:

  - specify the position argument as "dodge"

```{r echo = T}
diamond.bp2 + geom_bar(position = "dodge")
```

*************************

## Part 2
### Histograms - Diamonds data

Use `geom_histogram` to create the histograms you see below.

#### Plot a

Details:

  - binwidth used 0.15

```{r echo = T}
ggplot(data = diamonds, mapping = aes(x = carat, fill = cut)) + geom_histogram(binwidth = 0.15)
```

#### Plot b

Details:

  - binwidth used: 1000
  - colors used: "lightblue", "darkgreen", "darkorange", "darkred"
  - red line represents the mean price
  - orange line represents the median price
  - create vertical lines with `geom_vline`

```{r echo = T}
ggplot(data = diamonds, mapping = aes(x = price)) + 
  geom_histogram(binwidth = 1000, fill = "lightblue", color = "darkgreen") + 
  geom_vline(xintercept = mean(diamonds$price), color = "darkred") +    
  geom_vline(xintercept = median(diamonds$price), color = "darkorange") + 
  labs(title = "Diamond Prices", caption = "Skewed right distribution")
```


*************************

## Part 3
### Box plots - Diamonds data

Use `geom_boxplot` to create box plots for the plots you see below.

#### Plot a

Details:

  - colors used: "darkgrey", "purple"
  - use `x = reorder(cut, price, FUN = median)` to arrange box plots 
    by median value
  - use `coord_flip` for horizontal box plots

```{r echo = T}
ggplot(data = diamonds, 
       mapping = aes(x = reorder(cut, price, FUN = median), y = price)) +
  geom_boxplot(color = "darkgrey", fill = "purple") + coord_flip() +
  labs(x = "cut")
```

#### Plot b

Details:

  - colors used: "black", "purple"

```{r echo = T}
ggplot(data = diamonds, mapping = aes(x = cut, y = price)) + 
  geom_boxplot(color = "black", fill = "purple") + coord_flip() +
  facet_wrap(~clarity, nrow = 2)
```

*************************

## Part 4
### Heat maps - Diamonds data

#### Plot a

Use `geom_raster` to create the heat maps for the plots you see below. The code provided below will create a new data frame that has the median price for each color and cut combination of the diamonds data frame.

```{r}
diamonds.median <- aggregate(formula = price ~ color + cut, 
                             data = diamonds, FUN = median)
str(diamonds.median)
```

Use the data frame `diamonds.median` to create the below heatmap with
`geom_raster`.

```{r echo = T}
ggplot(data = diamonds.median, mapping = aes(x = color, y = cut, fill = price)) +
  geom_raster() +
  labs(title = "Heatmap for the median price of diamonds")
```

```{r}
diamonds.median2 <- aggregate(formula = price ~ color + cut + clarity, 
                              data = diamonds, FUN = median)
str(diamonds.median2)
```

Use the data frame `diamonds.median2` to create the below faceted heat maps
with `geom_raster`.

```{r echo = T}
ggplot(data = diamonds.median2, mapping = aes(x = color, y = cut, fill = price)) +
  geom_raster() + 
  facet_wrap(~clarity, nrow = 2) + 
  labs(title = "Heatmap for the median price of diamonds")
```

## Part 5
### Density plot

Below is an example of a density plot. Run the ggplot part line by line to see
what is happening as each layer is added.

```{r}
# generate 1000000 random variables from two normal distributions
x <- rnorm(1000000, mean = 0, sd = 1)
y <- rnorm(1000000, mean = 3, sd = 2)

# create a data frame
norm_df <- data.frame(value = c(x, y), norm = rep(c("x", "y"), each = 1000000))

# plot the densities
ggplot(data = norm_df, aes(x = value, fill = norm)) + 
  geom_density(alpha = 0.50) +
  labs(x = "Value", y = "Density", 
       title = "Normal probability density functions")+
  theme_bw() +
  scale_fill_discrete(name = "",
                      breaks = c("x", "y"),
                      labels = c("Normal(0, 1)", "Normal(3, 2)"))
```

