---
title: 'STT 301: More graphics with ggplot2'
author: "Rellika Kisyula"
date: "10/8/2018"
output: html_document
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
**Learning objectives:**

1. ggplot function
2. geom_bar
3. geom_histogram
4. geom_boxplot
5. geom_raster
6. geom_density
The package ggplot2 has many options and capabilities so you will probably find the following ggplot2 help resources useful:

+ http://docs.ggplot2.org/current/
+ the graphics portion of the R cookbook site at http://www.cookbook-r.com/Graphs/
+ http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/
+ RStudio “Data Visualization with ggplot2” cheatsheet
+ Google

# Part 1
## Bar plots - Diamonds data
The ggplot2 package comes with a data set called diamonds. Let’s look at it below. To obtain further details type ?diamonds in your console window.

```{r}
library(ggplot2)
str(diamonds)
```

Last time you created scatter plots such as the one below by making use of `geom_point`.

```{r}
diamond.plot <- ggplot(data=diamonds, aes(x=carat, y=price, colour = color))
diamond.plot + geom_point() + facet_wrap(~cut)
```

Use `geom_bar` to create the bar plots you see below.
**Plot a**
```{r}
ggplot(diamonds, aes(x = cut)) + geom_bar()
```

**Plot b**
```{r}
ggplot(diamonds, aes(x = cut, fill = clarity)) + geom_bar()
```
**Plot c**
```{r}
ggplot(diamonds, aes(x = cut, fill = clarity)) +
  geom_bar(position = "dodge")
```

# Part 2
## Histograms - Diamonds data
Use `geom_histogram` to create the histograms you see below.

**Plot a**
```{r}
ggplot(diamonds, aes(x = carat, fill = cut)) +
geom_histogram(binwidth = 0.15 )
```

**Plot b**
Details:

+ binwidth used: 1000
+ colors used: “lightblue”, “darkgreen”, “darkorange”, “darkred”
+ red line represents the mean price
+ orange line represents the median price
+ create vertical lines with `geom_vline`

```{r}
ggplot(diamonds, aes(x = price)) +
  geom_histogram(binwidth = 1000, fill = "lightblue", color ="darkgreen" ) +
  geom_vline(aes(xintercept = mean(price)), color="darkred") +
  geom_vline(aes(xintercept = median(price)), color="darkorange")
```

# Part 3
##Box plots - Diamonds data
Use `geom_boxplot` to create box plots for the plots you see below.

**Plot a**
Details:

+ colors used: “darkgrey”, “purple”
+ use `x = reorder(cut, price, FUN = median)` to arrange box plots by median value
+ use `coord_fli`p for horizontal box plots

```{r}
ggplot(data = diamonds,
    aes(x = reorder(cut, price, FUN = median), y = price)) +
  geom_boxplot(color = "darkgrey", fill = "purple") +
  coord_flip() +
  labs(x = "count", y="price")
```

Plot b
Details:

+ colors used: “black”, “purple”

```{r}
ggplot(data = diamonds,
    aes(x = reorder(cut, price, FUN = median), y = price)) +
  geom_boxplot(color = "darkgrey", fill = "purple") +
  coord_flip() +
  labs(x = "cut", y="price") +
  facet_wrap(~factor(clarity), nrow = 2)
```


# Part 4
## Heat maps - Diamonds data
**Plot a**
Use `geom_raster` to create the heat maps for the plots you see below. The code provided below will create a new data frame that has the median price for each color and cut combination of the diamonds data frame.

```{r}
diamonds.median <- aggregate(formula = price ~ color + cut, 
                             data = diamonds, FUN = median)
str(diamonds.median)
```

Use the data frame `diamonds.median` to create the below heatmap with geom_raster.
```{r}
ggplot(diamonds.median, aes(color, cut)) +
  geom_raster(aes(fill = price)) +
  labs(title = "Heatmap for the median price of the diamonds")
```

```{r}
diamonds.median2 <- aggregate(formula = price ~ color + cut + clarity, 
                              data = diamonds, FUN = median)
str(diamonds.median2)

```

Use the data frame `diamonds.median2` to create the below faceted heat maps with `geom_raster`.

```{r}
ggplot(diamonds.median2, aes(color, cut)) +
  geom_raster(aes(fill = price)) +
  facet_wrap(~factor(clarity), nrow = 2) +
  labs(title = "Heatmap for the median price of the diamonds")
```

# Part 5
## Density plot
Below is an example of a density plot. Run the ggplot part line by line to see what is happening as each layer is added.

```{r}
# generate 1000000 random variables from two normal distributions
x <- rnorm(1000000, mean = 0, sd = 1)
y <- rnorm(1000000, mean = 3, sd = 2)

# create a data frame
norm_df <- data.frame(value = c(x, y), norm = rep(c("x", "y"), each = 1000000))

# plot the densities
ggplot(data = norm_df, aes(x = value, fill = norm)) + 
  geom_density(alpha = 0.50) +
  labs(x = "Value", y = "Density", 
       title = "Normal probability density functions")+
  theme_bw() +
  scale_fill_discrete(name = "",
                      breaks = c("x", "y"),
                      labels = c("Normal(0, 1)", "Normal(3, 2)"))

```
