---
title: "STT 301: Web Scraping"
author: "Shawn Santo"
date: "November 26, 2018"
output: html_document
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(fig.path='Figs/', warning=FALSE, message=FALSE, dev = "png", comment = NA)
```

## Introduction

Learning Objectives

  - web scraping
  - SelectorGadget
  - rvest
  
*************************

```{r}
library(rvest)
library(tidyverse)
```

## Kenpom

### Scrape

  1. We will scrape some college basketball statistics using SelectorGadget. 
  The advanced metrics are available at: [https://kenpom.com/](https://kenpom.com/).
  Use SelectorGadget along with the following functions in the `rvest` 
  package: `read_html`, `html_nodes`, `html_text` to scrape the large table on 
  the main page. Only scrape the data and not the column headers. Save the 
  result as `kenpom.text`. The length of kenpom.text should be 7413.

```{r}
kenpom.url <- "https://kenpom.com/"
kenpom.text <- read_html(kenpom.url)

kenpom.text <- kenpom.text %>% 
    html_nodes(css = "table") %>%
    html_table(fill = TRUE)
```

### Organize the data

  1. We will organize the 353 NCAAB teams into a 353 by 21 matrix. We will then 
  remove the odd columns 7 through 21. These columns represent the rank 
  corresponding to the adjacent statistic. If you look on 
  [https://kenpom.com/](https://kenpom.com/) you will see small numbers next to 
  some of the statistics such as AdjO, AdjD, etc. This is what we are removing. 
  Once you have completed the above R chunk you may remove the chunk 
  option `eval = FALSE` below.

```{r eval = FALSE}
kenpom.mat <- matrix(kenpom.text, 
                     nrow = length(kenpom.text)/21, 
                     ncol = 21, 
                     byrow = TRUE)

kenpom.mat <- kenpom.mat[,-c(seq(7, 21, by = 2))]
kenpom.mat
```

  2. Create a data frame of `kenpom.mat` and save the result as `kenpom`. 
  Be sure to specify `stringsAsFactors = FALSE`. Assign the column names 
  as are given below. Look at the structure of `kenpom`. You can also 
  turn `kenpom` into a tibble if you like. Once you have completed the 
  above R chunk you may remove the chunk option `eval = FALSE` below.

```{r eval = FALSE}

names(kenpom) <- c("rank", "team", "conf", "w_l", "adjem", "adjo", 
                   "adjd", "adjt", "luck", "adjem_sos", "oppo_sos",
                   "oppd_sos", "adjem_nc")

```

### Manipulate the data

  1. Use the `separate` function to turn the `w_l` column 
     into two columns: "wins", "losses".
  2. Change the variable type of `rank` and `wins:adjem_nc` from 
     character to double (numeric).
  3. Add an additional variable to `kenpom` called `win_percentage`. 
     `win_percentage` is the number of wins divided by total games 
     played (wins + losses).

  
```{r}

```

### Team comparison

  1. Create a 353 by 353 matrix that represents the difference between each 
     team's `adjem` metric for all 353 teams. `adjem` is the difference between 
     a team's offensive and defensive efficiency. Save the result as 
     `kenpom.adjem.mat`. The diagonal of the resulting matrix should be 0. 
     The `outer` function in R will allow you to do this in one line of code. 
     I give a simple example below.
  2. Add the team names to the rows and columns of `kenpom.adjem.mat`. 
  3. On Tuesday, 11/27, Michigan State plays Louisville. Use your matrix 
     `kenpom.adjem.mat` to find the difference in `adjem` between 
     "Michigan St." and "Louisville".
  
```{r}
# an example with outer function
outer(X = c(1:4), Y = c(1:4), FUN = "-")

```

For details on some of the advanced metrics see: [https://kenpom.com/blog/ratings-methodology-update/](https://kenpom.com/blog/ratings-methodology-update/).

*************************
## IMDb

### Movie data frame

  1. Expand on the IMDb top 100 working example of the lecture slides. 
  Scrape the movie run time, gross dollars, and the MPAA film rating. 
  Add these variables to the `imdb` data frame created in the working 
  example of the lecture slides.
  2. Use `ggplot` to create a graphic of your choice based on the 
  `imdb` data frame. Note, you may have to clean the data prior to using 
  `ggplot` depending on what variables you want to use.
  
