---
title: "STT 301: Monty Hall Simulation"
author: "Shawn Santo"
date: "October 24, 2018"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, comment = NA)
```

## Introduction

Learning Objectives

  - Simulation
  
*************************

### Game set-up and assumptions

"Suppose you're on a game show, and you're given the choice of three doors: Behind one door is a car; behind the others, goats. You pick a door, say No. 1, and the host, who knows what's behind the doors, opens another door, say No. 3, which has a goat. He then says to you, "Do you want to pick door No. 2?" Is it to your advantage to switch your choice?"

We will investigate the above decision of whether to switch or not to switch.

Assumptions:

  1. The host will always open a door not picked by the contestant.
  2. The host will always open a door which reveals a goat (i.e. not a car).
  3. The host will always offer the contestant the chance to switch to another door.
  4. The door behind which the car is placed is chosen at random.
  5. The door initially chosen by the contestant is chosen at random.

### Single game result

Write a function `monty_hall_one_play` that takes no arguments and returns a numeric vector with each element representing the following:

  - the first element is the door behind which the car is hidden (selected randomly from `1:3`)
  - the second element is the door initially selected by the contestant (selected randomly from `1:3`)
  - the third elment is the door revealed by the host
    - if the contestant selected the door with the car, the host randomly selects one of the two remaining doors
    - if the contestant selected a door different from the door with the car, the host selects the door that is not hiding the car

```{r echo = T}
monty_hall_one_play <- function(){
  
  door.options <- 1:3
  
  car.door <- sample(door.options, size = 1, replace = FALSE)
  user.door <- sample(door.options, size = 1, replace = FALSE)
  
  if (car.door == user.door){
    host.door <- sample(door.options[-car.door], size = 1, replace = FALSE)
  }else{
    host.door <- door.options[-c(car.door, user.door)]
  }
  
  return(c(car.door, user.door, host.door))
  
}
```

```{r}
monty_hall_one_play()
```

### Repeated plays

To evaluate the stay and switch strategies we must repeat the game many times to determine the proportion of wins based on each strategy. To do this, write another function called `monty_hall` which has one argument, `iter`, the number of iterations, and which returns a data frame with `iter` rows and five columns. This function should call the `monty_hall_one_play` function in the  body of code. You will probably want to write some type of loop, but maybe you can be clever and use one of the `apply` functions. Try the loop technique first.

The data frame columns that the function `monty_hall` should return are:

  - `car_door`, gives the door behind which the car is hidden.
  - `cont_door`, gives the door initially selected by the contestant.
  - `host_door`, gives the door revealed by the host.
  - `switch`, result of switch strategy; 1 is win, 0 is loss.
  - `stay`, result of stay strategy; 1 is win, 0 is loss.

```{r echo = T}
monty_hall <- function(iter){
  
  monty.hall.df <- data.frame(car_door = numeric(iter), cont_door = numeric(iter), host_door = numeric(iter), switch = numeric(iter), stay = numeric(iter))
  
  for (i in 1:iter){
    monty.hall.df[i,1:3] <- monty_hall_one_play()
  }
  
  monty.hall.df["switch"] <- as.numeric(monty.hall.df["car_door"] !=  monty.hall.df["cont_door"])
  monty.hall.df["stay"] <- as.numeric(!monty.hall.df["switch"])
  
  return(monty.hall.df)
  
}
```

We can look at the results for a small number of iterations 
to see if it was coded correctly.

```{r}
monty_hall(iter = 5)
```

### Strategy evaluation

Conduct a simulation with 10,000 iterations. Save the result as `monty.hall.sim`. Use the `apply` function to compute the mean of each column. You can now see a simulated win probability for each strategy.

```{r}
monty.hall.sim <- monty_hall(iter = 10000)
```

```{r echo = T, eval = T}
apply(monty.hall.sim, 2, mean)
```

Which strategy appears to be the best based on your simulation? You can verify these results using Bayes' Formula or a tree diagram. 

For some interesting reaction to the problem: http://marilynvossavant.com/game-show-problem/.